\chapter{Design Choices and Extensions}

Several design choices in \clisp{} are left to the individual
implementation, and some essential parts of the programming environment
are left undefined.  This chapter discusses the most important design
choices and extensions.

\section{Data Types}

\subsection{Integers}

The \tindexed{fixnum} type is equivalent to \code{(signed-byte 30)}.
Integers outside this range are represented as a \tindexed{bignum} or
a word integer (\pxlref{word-integers}.)  Almost all integers that
appear in programs can be represented as a \code{fixnum}, so integer
number consing is rare.


\subsection{Floats}
\label{ieee-float}

\cmucl{} supports three floating point formats:
\tindexed{single-float}, \tindexed{double-float} and
\tindexed{double-double-float}.  The first two are implemented with
IEEE single and double float arithmetic, respectively.  The last is an
extension; \pxlref{extended-float} for more information.
\code{short-float} is a synonym for \code{single-float}, and
\code{long-float} is a synonym for \code{double-float}.  The initial
value of \vindexed{read-default-float-format} is \code{single-float}.

Both \code{single-float} and \code{double-float} are represented with
a pointer descriptor, so float operations can cause number consing.
Number consing is greatly reduced if programs are written to allow the
use of non-descriptor representations (\pxlref{numeric-types}.)


\subsubsection{IEEE Special Values}

\cmucl{} supports the IEEE infinity and NaN special values.  These
non-numeric values will only be generated when trapping is disabled
for some floating point exception (\pxlref{float-traps}), so users of
the default configuration need not concern themselves with special
values.

\begin{defconst}{extensions:}{short-float-positive-infinity}
  \defconstx[extensions:]{short-float-negative-infinity}
  \defconstx[extensions:]{single-float-positive-infinity}
  \defconstx[extensions:]{single-float-negative-infinity}
  \defconstx[extensions:]{double-float-positive-infinity}
  \defconstx[extensions:]{double-float-negative-infinity}
  \defconstx[extensions:]{long-float-positive-infinity}
  \defconstx[extensions:]{long-float-negative-infinity}
  
  The values of these constants are the IEEE positive and negative
  infinity objects for each float format.
\end{defconst}

\begin{defun}{extensions:}{float-infinity-p}{\args{\var{x}}}
  
  This function returns true if \var{x} is an IEEE float infinity (of
  either sign.)  \var{x} must be a float.
\end{defun}

\begin{defun}{extensions:}{float-nan-p}{\args{\var{x}}}
  \defunx[extensions:]{float-signaling-nan-p}{\args{\var{x}}}
  \defunx[extensions:]{float-trapping-nan-p}{\args{\var{x}}}
  
  \code{float-nan-p} returns true if \var{x} is an IEEE NaN (Not A
  Number) object.  \code{float-signaling-nan-p} returns true only if
  \var{x} is a trapping NaN.  With either function, \var{x} must be a
  float. \code{float-trapping-nan-p} is the former name of
  \code{float-signaling-nan-p} and is deprecated.
\end{defun}

\subsubsection{Negative Zero}

The IEEE float format provides for distinct positive and negative
zeros.  To test the sign on zero (or any other float), use the
\clisp{} \findexed{float-sign} function.  Negative zero prints as
\code{-0.0f0} or \code{-0.0d0}.

\subsubsection{Denormalized Floats}

\cmucl{} supports IEEE denormalized floats.  Denormalized floats
provide a mechanism for gradual underflow.  The \clisp{}
\findexed{float-precision} function returns the actual precision of a
denormalized float, which will be less than \findexed{float-digits}.
Note that in order to generate (or even print) denormalized floats,
trapping must be disabled for the underflow exception
(\pxlref{float-traps}.)  The \clisp{}
\w{\code{least-positive-}\var{format}-\code{float}} constants are
denormalized.

\begin{defun}{extensions:}{float-denormalized-p}{\args{\var{x}}}
  
  This function returns true if \var{x} is a denormalized float.
  \var{x} must be a float.
\end{defun}


\subsubsection{Floating Point Exceptions}
\label{float-traps}

The IEEE floating point standard defines several exceptions that occur
when the result of a floating point operation is unclear or
undesirable.  Exceptions can be ignored, in which case some default
action is taken, such as returning a special value.  When trapping is
enabled for an exception, a error is signalled whenever that exception
occurs.  These are the possible floating point exceptions:
\begin{Lentry}
  
\item[\kwd{underflow}] This exception occurs when the result of an
  operation is too small to be represented as a normalized float in
  its format.  If trapping is enabled, the
  \tindexed{floating-point-underflow} condition is signalled.
  Otherwise, the operation results in a denormalized float or zero.
  
\item[\kwd{overflow}] This exception occurs when the result of an
  operation is too large to be represented as a float in its format.
  If trapping is enabled, the \tindexed{floating-point-overflow}
  exception is signalled.  Otherwise, the operation results in the
  appropriate infinity.
  
\item[\kwd{inexact}] This exception occurs when the result of a
  floating point operation is not exact, i.e. the result was rounded.
  If trapping is enabled, the \code{extensions:floating-point-inexact}
  condition is signalled.  Otherwise, the rounded result is returned.
  
\item[\kwd{invalid}] This exception occurs when the result of an
  operation is ill-defined, such as \code{\w{(/ 0.0 0.0)}}.  If
  trapping is enabled, the \code{extensions:floating-point-invalid}
  condition is signalled.  Otherwise, a quiet NaN is returned.
  
\item[\kwd{divide-by-zero}] This exception occurs when a float is
  divided by zero.  If trapping is enabled, the
  \tindexed{divide-by-zero} condition is signalled.  Otherwise, the
  appropriate infinity is returned.
\end{Lentry}

\subsubsection{Floating Point Rounding Mode}
\label{float-rounding-modes}

IEEE floating point specifies four possible rounding modes:
\begin{Lentry}
  
\item[\kwd{nearest}] In this mode, the inexact results are rounded to
  the nearer of the two possible result values.  If the neither
  possibility is nearer, then the even alternative is chosen.  This
  form of rounding is also called ``round to even'', and is the form
  of rounding specified for the \clisp{} \findexed{round} function.
  
\item[\kwd{positive-infinity}] This mode rounds inexact results to the
  possible value closer to positive infinity.  This is analogous to
  the \clisp{} \findexed{ceiling} function.
  
\item[\kwd{negative-infinity}] This mode rounds inexact results to the
  possible value closer to negative infinity.  This is analogous to
  the \clisp{} \findexed{floor} function.
  
\item[\kwd{zero}] This mode rounds inexact results to the possible
  value closer to zero.  This is analogous to the \clisp{}
  \findexed{truncate} function.
\end{Lentry}

Warning: Although the rounding mode can be changed with
\code{set-floating-point-modes}, use of any value other than the
default (\kwd{nearest}) can cause unusual behavior, since it will
affect rounding done by \llisp{} system code as well as rounding in
user code.  In particular, the unary \code{round} function will stop
doing round-to-nearest on floats, and instead do the selected form of
rounding.

%% \subsubsection{Precision Control}
%% \label{precision-control}
%% 
%% The floating-point unit for the Intel IA-32 architecture supports a
%% precision control mechanism.  The floating-point unit consists of an
%% IEEE extended double-float unit and all operations are always done
%% using his format, and this includes rounding.  However, by setting the
%% precision control mode, the user can control how rounding is done for
%% each basic arithmetic operation like addition, subtraction,
%% multiplication, and division.  The extra instructions for
%% trigonometric, exponential, and logarithmic operations are not
%% affected.  We refer the reader to Intel documentation for more
%% information. 
%% 
%% The possible modes are:
%% \begin{Lentry}
%%   
%% \item[\kwd{24-bit}] In this mode, all basic arithmetic operations like
%%   addition, subtraction, multiplication, and division, are rounded
%%   after each operation as if both the operands were IEEE single
%%   precision numbers.  
%%   
%% \item[\kwd{53-bit}] In this mode, rounding is performed as if the
%%   operands and results were IEEE double precision numbers.
%%   
%% \item[\kwd{64-bit}] In this mode, the default, rounding is performed
%%   on the full IEEE extended double precision format.
%%   
%% \end{Lentry}
%% 
%% \subsubsection{Warning:}
%% 
%% Although the precision mode can be changed with
%% \code{set-floating-point-modes}, use of anything other than
%% \kwd{64-bit} or \kwd{53-bit} can cause unexpected results, especially
%% if external functions or libraries are called.  A setting of
%% \kwd{64-bit} also causes \code{(= 1d0 (+ 1d0 double-float-epsilon))}
%% to return \true{} instead of \false.
%% 
%% 
\subsubsection{Accessing the Floating Point Modes}

These functions can be used to modify or read the floating point modes:

\begin{defun}{extensions:}{set-floating-point-modes}{%
    \keys{\kwd{traps} \kwd{rounding-mode}}
    \morekeys{\kwd{fast-mode} \kwd{accrued-exceptions}}
    \yetmorekeys{\kwd{current-exceptions}}}
  \defunx[extensions:]{get-floating-point-modes}{}
  
  The keyword arguments to \code{set-floating-point-modes} set various
  modes controlling how floating point arithmetic is done:
  \begin{Lentry}
  
  \item[\kwd{traps}] A list of the exception conditions that should
    cause traps.  Possible exceptions are \kwd{underflow},
    \kwd{overflow}, \kwd{inexact}, \kwd{invalid} and
    \kwd{divide-by-zero}.  Initially all traps except \kwd{inexact}
    are enabled.  \xlref{float-traps}.
    
  \item[\kwd{rounding-mode}] The rounding mode to use when the result
    is not exact. Possible values are \kwd{nearest},
    \kwd{positive-infinity}, \kwd{negative-infinity} and \kwd{zero}.
    Initially, the rounding mode is \kwd{nearest}. See the warning in
    section \ref{float-rounding-modes} about use of other rounding
    modes.
  
  \item[\kwd{current-exceptions}, \kwd{accrued-exceptions}] Lists of
    exception keywords used to set the exception flags.  The
    \var{current-exceptions} are the exceptions for the previous
    operation, so setting it is not very useful.  The
    \var{accrued-exceptions} are a cumulative record of the exceptions
    that occurred since the last time these flags were cleared.
    Specifying \code{()} will clear any accrued exceptions.
  
  \item[\kwd{fast-mode}] Set the hardware's ``fast mode'' flag, if
    any.  When set, IEEE conformance or debuggability may be impaired.
    Some machines may not have this feature, in which case the value
    is always \false.  Sparc platforms support a fast mode where
    denormal numbers are silently truncated to zero.
  \end{Lentry}
  If a keyword argument is not supplied, then the associated state is
  not changed.
  
  \code{get-floating-point-modes} returns a list representing the
  state of the floating point modes.  The list is in the same format
  as the keyword arguments to \code{set-floating-point-modes}, so
  \code{apply} could be used with \code{set-floating-point-modes} to
  restore the modes in effect at the time of the call to
  \code{get-floating-point-modes}.
\end{defun}

To make handling control of floating-point exceptions, the following
macro is useful.

\begin{defmac}{ext:}{with-float-traps-masked}{\var{traps} \ampbody\ \var{body}}
  \code{body} is executed with the selected floating-point exceptions
  given by \code{traps} masked out (disabled).  \code{traps} should be
  a list of possible floating-point exceptions that should be ignored.
  Possible values are \kwd{underflow}, \kwd{overflow}, \kwd{inexact},
  \kwd{invalid} and \kwd{divide-by-zero}.
  
  This is equivalent to saving the current traps from
  \code{get-floating-point-modes}, setting the floating-point modes to
  the desired exceptions, running the \code{body}, and restoring the
  saved floating-point modes.  The advantage of this macro is that it
  causes less consing to occur.

  Some points about the with-float-traps-masked:

  \begin{itemize}
  \item Two approaches are available for detecting FP exceptions:
    \begin{enumerate}
    \item enabling the traps and handling the exceptions
    \item disabling the traps and either handling the return values or
      checking the accrued exceptions.
    \end{enumerate}
    Of these the latter is the most portable because on the alpha port
    it is not possible to enable some traps at run-time.
    
  \item To assist the checking of the exceptions within the body any
    accrued exceptions matching the given traps are cleared at the
    start of the body when the traps are masked.
    
  \item To allow the macros to be nested these accrued exceptions are
    restored at the end of the body to their values at the start of
    the body. Thus any exceptions that occurred within the body will
    not affect the accrued exceptions outside the macro.
    
  \item Note that only the given exceptions are restored at the end of
    the body so other exception will be visible in the accrued
    exceptions outside the body.
    
  \item On the x86, setting the accrued exceptions of an unmasked
    exception would cause a FP trap. The macro behaviour of restoring
    the accrued exceptions ensures than if an accrued exception is
    initially not flagged and occurs within the body it will be
    restored/cleared at the exit of the body and thus not cause a
    trap.
    
  \item On the x86, and, perhaps, the hppa, the FP exceptions may be
    delivered at the next FP instruction which requires a FP
    \code{wait} instruction (\code{x86::float-wait}) if using the lisp
    conditions to catch trap within a \code{handler-bind}.  The
    \code{handler-bind} macro does the right thing and inserts a
    float-wait (at the end of its body on the x86).  The masking and
    noting of exceptions is also safe here.
    
  \item The setting of the FP flags uses the
    \code{(floating-point-modes)} and the \code{(set
      (floating-point-modes)\ldots)} VOPs. These VOPs blindly update
    the flags which may include other state.  We assume this state
    hasn't changed in between getting and setting the state. For
    example, if you used the FP unit between the above calls, the
    state may be incorrectly restored! The
    \code{with-float-traps-masked} macro keeps the intervening code to
    a minimum and uses only integer operations.
    %% Safe byte-compiled?
    %% Perhaps the VOPs (x86) should be smarter and only update some of
    %% the flags, the trap masks and exceptions?
  \end{itemize}

\end{defmac}

\subsection{Extended Floats}
\label{extended-float}

\cmucl{} also has an extension to support \code{double-double-float}
type.  This float format provides extended precision of about 31
decimal digits, with the same exponent range as \code{double-float}.
It is completely integrated into \cmucl{}, and can be used just like
any other floating-point object, including arrays, complex
\code{double-double-float}'s, and special functions.  With appropriate
declarations, no boxing is needed, just like \code{single-float} and
\code{double-float}. 

The exponent marker for a double-double float number is ``W'', so
``1.234w0'' is a double-double float number.


Note that there are a few shortcomings with
\code{double-double-float}'s:
\begin{itemize}
 \item There are no equivalents to \code{most-positive-double-float},
   \code{double-float-positive-infinity}, \textit{etc}.  This is because
   these are not really well defined for \code{double-double-float}'s.
 \item Underflow and overflow may be prematurely signaled.  This is
   due to how \code{double-double-float}'s are implemented.
 \item Basic arithmetic operations are inlined, so the code size is
   fairly large.
 \item \code{double-double-float} arithmetic is quite a bit slower
   than \code{double-float} since there is no hardware support for
   this type.
 \item The constant \code{pi} is still a \code{double-float} instead
   of a \code{double-double-float}.  Use \code{ext:dd-pi} if you
   want a \code{double-double-float} value for $\pi$.
\end{itemize}

\begin{deftp}{float}{extensions:double-double-float}{}
  The \code{double-double-float} type.  It is in the \code{EXTENSIONS}
  package.
\end{deftp}

\begin{defconst}{extensions:}{dd-pi}
  A \code{double-double-float} approximation to $\pi$.
\end{defconst}

\subsection{Characters}

\cmucl{} implements characters according to \cltltwo{}. The
main difference from the first version is that character bits and font
have been eliminated, and the names of the types have been changed.
\tindexed{base-character} is the new equivalent of the old
\tindexed{string-char}. In this implementation, all characters are
base characters (there are no extended characters.) Character codes
range between \code{0} and \code{255}, using the ASCII encoding.
Table~\ref{tbl:chars}~\vpageref{tbl:chars} shows characters recognized
by \cmucl.

\begin{table}[tbhp]
  \begin{center}
    \begin{tabular}{|c|c|l|l|l|l|}
      \hline
      \multicolumn{2}{|c|}{ASCII} & \multicolumn{1}{|c}{Lisp} &
      \multicolumn{3}{|c|}{} \\
      \cline{1-2}
      Name & Code & \multicolumn{1}{|c|}{Name} & \multicolumn{3}{|c|}{\raisebox{1.5ex}{Alternatives}}\\
      \hline
      \hline
      \code{nul} & 0 & \code{\#\back{NULL}} & \code{\#\back{NUL}} & &\\
      \code{bel} & 7 & \code{\#\back{BELL}} & & &\\
      \code{bs} &  8 & \code{\#\back{BACKSPACE}} & \code{\#\back{BS}} & &\\
      \code{tab} & 9 & \code{\#\back{TAB}} & & &\\
      \code{lf} & 10 & \code{\#\back{NEWLINE}} & \code{\#\back{NL}} & \code{\#\back{LINEFEED}} & \code{\#\back{LF}}\\
      \code{ff} & 11 & \code{\#\back{VT}} & \code{\#\back{PAGE}} & \code{\#\back{FORM}} &\\
      \code{cr} & 13 & \code{\#\back{RETURN}} & \code{\#\back{CR}} & &\\
      \code{esc} & 27 & \code{\#\back{ESCAPE}} & \code{\#\back{ESC}} & \code{\#\back{ALTMODE}} & \code{\#\back{ALT}}\\
      \code{sp} & 32 & \code{\#\back{SPACE}} & \code{\#\back{SP}} & &\\
      \code{del} & 127 & \code{\#\back{DELETE}} & \code{\#\back{RUBOUT}} & &\\
      \hline
    \end{tabular}
    \caption{Characters recognized by \cmucl}
    \label{tbl:chars}
  \end{center}
\end{table}


\subsection{Array Initialization}

If no \kwd{initial-value} is specified, arrays are initialized to zero.


\subsection{Hash tables}

The \tindexed{hash-tables} defined by \clisp{} have limited utility because they
are limited to testing their keys using the equality predicates
provided by (pre-CLOS) \clisp{}.  \cmucl{} overcomes this limitation
by allowing its users to specify new hash table tests and hashing
methods.  The hashing method must also be specified, since the
compiler is unable to determine a good hashing function for an
arbitrary equality (equivalence) predicate.

\begin{defun}{extensions:}{define-hash-table-test}%
  {\args{\var{hash-table-test-name} \var{test-function} \var{hash-function}}}
      
      The \var{hash-table-test-name} must be a symbol.
      % I just assumed the above. [2002/10/10:rpg]
      The \var{test-function} takes two objects and returns true
      iff they are the same.  The \var{hash-function} takes one object and
      returns two values: the (positive fixnum) hash value and true if
      the hashing depends on pointer values and will have to be redone
      if the object moves.
      
      To create a hash-table using this new ``test'' (really, a
      test/hash-function pair), use
      \code{(\index[funs]{make-hash-table}make-hash-table :test
        \var{hash-table-test-name} \ldots)}.

      Note that it is the \var{hash-table-test-name} that will be
      returned by the function \findexed{hash-table-test}, when applied to
      a hash-table created using this function.

      This function updates \vindexed{hash-table-tests}, which is now
      internal.  
\end{defun}

\cmucl{} also supports a number of weak hash tables.  These weak
tables are created using the \kwd{weak-p} argument to
\code{make-hash-table}.  Normally, a reference to an object as either
the key or value of the hash-table will prevent that object from being
garbage-collected.  However, in a weak table, if the only reference is
the hash-table, the object can be collected.

The possible values for \kwd{weak-p} are listed below.  An entry in
the table remains if the condition holds
\begin{Lentry}
\item[\kwd{key}] The key is referenced elsewhere
\item[\kwd{value}] The value is referenced elsewhere
\item[\kwd{key-and-value}] Both the key and value are referenced elsewhere
\item[\kwd{key-or-value}] Either the key or value are referenced elsewhere
\item[T] For backward compatibility, this means the same as \kwd{key}.
\end{Lentry}
If the condition does not hold, the object can be removed from the
hash table.  

Weak hash tables can only be created if the test is \code{eq} or
\code{eql}.  An error is signaled if this is not the case.

\begin{defun}{}{make-hash-table}%
  {\args{\keys{\kwd{test} \kwd{size} \kwd{rehash-size} \kwd{rehash-threshold} \kwd{weak-p}}}}
  Creates a hash-table with the specified properties.
\end{defun}
\section{Default Interrupts for Lisp}

\cmucl{} has several interrupt handlers defined when it starts up,
as follows:
\begin{Lentry}
  
\item[\code{SIGINT} (\ctrl{c})] causes Lisp to enter a break loop.
  This puts you into the debugger which allows you to look at the
  current state of the computation.  If you proceed from the break
  loop, the computation will proceed from where it was interrupted.
  
\item[\code{SIGQUIT} (\ctrl{L})] causes Lisp to do a throw to the
  top-level.  This causes the current computation to be aborted, and
  control returned to the top-level read-eval-print loop.
  
\item[\code{SIGTSTP} (\ctrl{z})] causes Lisp to suspend execution and
  return to the Unix shell.  If control is returned to Lisp, the
  computation will proceed from where it was interrupted.
  
\item[\code{SIGILL}, \code{SIGBUS}, \code{SIGSEGV}, and \code{SIGFPE}]
  cause Lisp to signal an error.
\end{Lentry}
For keyboard interrupt signals, the standard interrupt character is in
parentheses.  Your \file{.login} may set up different interrupt
characters.  When a signal is generated, there may be some delay before
it is processed since Lisp cannot be interrupted safely in an arbitrary
place.  The computation will continue until a safe point is reached and
then the interrupt will be processed.  \xlref{signal-handlers} to define
your own signal handlers.


\section{Implementation-Specific Packages}

When \cmucl{} is first started up, the default package is the
\code{common-lisp-user} package. The \code{common-lisp-user} package
uses the \code{common-lisp} and \code{extensions} packages. The
symbols exported from these three packages can be referenced without
package qualifiers. This section describes packages which have
exported interfaces that may concern users. The numerous internal
packages which implement parts of the system are not described here.
Package nicknames are in parenthesis after the full name.

\begin{Lentry}
\item[\code{alien}, \code{c-call}] Export the features of the Alien
  foreign data structure facility (\pxlref{aliens}.)
  
\item[\code{pcl}] This package contains PCL (Portable CommonLoops),
  which is a portable implementation of CLOS (the Common Lisp Object
  System.)  This implements most (but not all) of the features in the
  CLOS chapter of \cltltwo{}.

\item[\code{clos-mop (mop)}] This package contains an implementation
  of the CLOS Metaobject Protocol, as per the book \textit{The Art of
  the Metaobject Protocol}.
  
\item[\code{debug}] The \code{debug} package contains the command-line
  oriented debugger.  It exports utility various functions and
  switches.
  
\item[\code{debug-internals}] The \code{debug-internals} package
  exports the primitives used to write debuggers.
  \xlref{debug-internals}.
  
\item[\code{extensions (ext)}] The \code{extensions} packages exports
  local extensions to \clisp{} that are documented in this manual.
  Examples include the \code{save-lisp} function and time parsing.
  
\item[\code{hemlock (ed)}] The \code{hemlock} package contains all the
  code to implement Hemlock commands.  The \code{hemlock} package
  currently exports no symbols.
  
\item[\code{hemlock-internals (hi)}] The \code{hemlock-internals}
  package contains code that implements low level primitives and
  exports those symbols used to write Hemlock commands.
  
\item[\code{keyword}] The \code{keyword} package contains keywords
  (e.g., \kwd{start}).  All symbols in the \code{keyword} package are
  exported and evaluate to themselves (i.e., the value of the symbol
  is the symbol itself).
  
\item[\code{profile}] The \code{profile} package exports a simple
  run-time profiling facility (\pxlref{profiling}).
  
\item[\code{common-lisp (cl)}] The \code{common-lisp} package
  exports all the symbols defined by \cltl{} and only those symbols.
  Strictly portable Lisp code will depend only on the symbols exported
  from the \code{common-lisp} package.
  
\item[\code{unix}] This package exports system call
  interfaces to Unix (\pxlref{unix-interface}).
  
\item[\code{system (sys)}] The \code{system} package contains
  functions and information necessary for system interfacing.  This
  package is used by the \code{lisp} package and exports several
  symbols that are necessary to interface to system code.
  
\item[\code{xlib}] The \code{xlib} package contains the Common Lisp X
  interface (CLX) to the X11 protocol.  This is mostly Lisp code with
  a couple of functions that are defined in C to connect to the
  server.
  
\item[\code{wire}] The \code{wire} package exports a remote procedure
  call facility (\pxlref{remote}).

\item[\code{stream}] The \code{stream} package exports the public
  interface to the simple-streams implementation (\pxlref{simple-streams}).

\item[\code{xref}] The \code{xref} package exports the public
  interface to the cross-referencing utility (\pxlref{xref}).

\end{Lentry}


\input{hierarchical-packages}

\input{package-locks}



\section{The Editor}

The \code{ed} function invokes the Hemlock editor which is described
in {\it Hemlock User's Manual} and {\it Hemlock Command Implementor's
Manual}. Most users at CMU prefer to use Hemlock's slave \llisp{}
mechanism which provides an interactive buffer for the
\code{read-eval-print} loop and editor commands for evaluating and
compiling text from a buffer into the slave \llisp.  Since the editor
runs in the \llisp, using slaves keeps users from trashing their
editor by developing in the same \llisp{} with \hemlock{}.


\section{Garbage Collection}

\cmucl{} uses either a stop-and-copy garbage collector or a
generational, mostly copying garbage collector.  Which collector is
available depends on the platform and the features of the platform.
The stop-and-copy GC is available on all RISC platforms.  The x86
platform supports a conservative stop-and-copy collector, which is now
rarely used, and a generational conservative collector.  On the Sparc
platform, both the stop-and-copy GC and the generational GC are
available, but the stop-and-copy GC is deprecated in favor of the
generational GC.  

The generational GC is available if \var{*features*} contains
\code{:gencgc}.

%% The stop-and-copy GC compacts the items in dynamic space every time it
%% runs. Most users cause the system to garbage collect (GC) frequently,
%% long before space is exhausted. With 16 or 24 megabytes of memory,
%% causing GC's more frequently on less garbage allows the system to GC
%% without much (if any) paging.

The following functions invoke the garbage collector or control whether
automatic garbage collection is in effect:

\begin{defun}[-cheney]{extensions:}{gc}{\args{\ampoptional{} \var{verbose-p}}}
  
  This function runs the garbage collector.  If
  \code{ext:*gc-verbose*} is non-\nil, then it invokes
  \code{ext:*gc-notify-before*} before GC'ing and
  \code{ext:*gc-notify-after*} afterwards.
  
  \code{verbose-p} indicates whether GC statistics are printed or
  not. 

\end{defun}

\begin{defun}{extensions:}{gc-off}{}
  
  This function inhibits automatic garbage collection.  After calling
  it, the system will not GC unless you call \code{ext:gc} or
  \code{ext:gc-on}.
\end{defun}

\begin{defun}{extensions:}{gc-on}{}
  
  This function reinstates automatic garbage collection.  If the
  system would have GC'ed while automatic GC was inhibited, then this
  will call \code{ext:gc}.
\end{defun}

\subsection{GC Parameters}

The following variables control the behavior of the garbage collector:

\begin{defvar}{extensions:}{bytes-consed-between-gcs}
  
  \cmucl{} automatically GC's whenever the amount of memory
  allocated to dynamic objects exceeds the value of an internal
  variable.  After each GC, the system sets this internal variable to
  the amount of dynamic space in use at that point plus the value of
  the variable \code{ext:*bytes-consed-between-gcs*}.  The default
  value is 2000000.
\end{defvar}

\begin{defvar}{extensions:}{gc-verbose}
  
  This variable controls whether \code{ext:gc} invokes the functions
  in \code{ext:*gc-notify-before*} and
  \code{ext:*gc-notify-after*}.  If \code{*gc-verbose*} is \nil,
  \code{ext:gc} foregoes printing any messages.  The default value is
  \code{T}.
\end{defvar}

\begin{defvar}{extensions:}{gc-notify-before}
  
  This variable's value is a function that should notify the user that
  the system is about to GC.  It takes one argument, the amount of
  dynamic space in use before the GC measured in bytes.  The default
  value of this variable is a function that prints a message similar
  to the following:
\begin{verbatim}
   [GC threshold exceeded with 2,107,124 bytes in use.  Commencing GC.]
\end{verbatim}
\end{defvar}

\begin{defvar}{extensions:}{gc-notify-after}
  
  This variable's value is a function that should notify the user when
  a GC finishes.  The function must take three arguments, the amount
  of dynamic spaced retained by the GC, the amount of dynamic space
  freed, and the new threshold which is the minimum amount of space in
  use before the next GC will occur.  All values are byte quantities.
  The default value of this variable is a function that prints a
  message similar to the following:
  \begin{verbatim}
    [GC completed with 25,680 bytes retained and 2,096,808 bytes freed.]
    [GC will next occur when at least 2,025,680 bytes are in use.]
  \end{verbatim}
\end{defvar}

Note that a garbage collection will not happen at exactly the new
threshold printed by the default \code{ext:*gc-notify-after*}
function.  The system periodically checks whether this threshold has
been exceeded, and only then does a garbage collection.

\begin{defvar}{extensions:}{gc-inhibit-hook}
  
  This variable's value is either a function of one argument or \nil.
  When the system has triggered an automatic GC, if this variable is a
  function, then the system calls the function with the amount of
  dynamic space currently in use (measured in bytes).  If the function
  returns \nil, then the GC occurs; otherwise, the system inhibits
  automatic GC as if you had called \code{ext:gc-off}.  The writer of
  this hook is responsible for knowing when automatic GC has been
  turned off and for calling or providing a way to call
  \code{ext:gc-on}.  The default value of this variable is \nil.
\end{defvar}

\begin{defvar}{extensions:}{before-gc-hooks}
  \defvarx[extensions:]{after-gc-hooks}
  
  These variables' values are lists of functions to call before or
  after any GC occurs.  The system provides these purely for
  side-effect, and the functions take no arguments.
\end{defvar}

\subsection{Generational GC}
Generational GC also supports some additional functions and variables
to control it.

\begin{defun}[-gencgc]{extensions:}{gc}{\args{\keys{\kwd{verbose} \kwd{gen} \kwd{full}}}}
  
  This function runs the garbage collector.  If
  \code{ext:*gc-verbose*} is non-\nil, then it invokes
  \code{ext:*gc-notify-before*} before GC'ing and
  \code{ext:*gc-notify-after*} afterwards.

  \begin{Lentry}
  \item[\code{verbose}] Print GC statistics if non-\code{NIL}.
  \item[\code{gen}] The number of generations to be collected.
  \item[\code{full}] If non-\code{NIL}, a full collection of all
    generations is performed.
  \end{Lentry}
\end{defun}

\begin{defun}{lisp::}{gencgc-stats}{\args{\var{generation}}}
  Returns statistics about the generation, as multiple values:
  \begin{enumerate}
  \item Bytes allocated in this generation
  \item The GC trigger for this generation.  When this many bytes have
    been allocated, a GC is started automatically.
  \item The number of bytes consed between GCs.
  \item The number of GCs that have been done on this generation.
    This is reset to zero when the generation is raised.
  \item The trigger age, which is the maximum number of GCs to perform
    before this generation is raised.
  \item The total number of bytes allocated to this generation.
  \item Average age of the objects in this generations.  The average
    age is the cumulative bytes allocated divided by current number of
    bytes allocated.
  \end{enumerate}
\end{defun}

\begin{defun}{lisp::}{set-gc-trigger}{\args{\var{gen} \var{trigger}}}
  Sets the GC trigger value for the specified generation.
\end{defun}

\begin{defun}{lisp::}{set-trigger-age}{\args{\var{gen} \var{trigger-age}}}
  Sets the GC trigger age for the specified generation.
\end{defun}

\begin{defun}{lisp::}{set-min-mem-age}{\args{\var{gen} \var{min-mem-age}}}
  Sets the minimum average memory age for the specified generation.
  If the computed memory age is below this, GC is not performed, which
  helps prevent a GC when a large number of new live objects have been
  added in which case a GC would usually be a waste of time.
\end{defun}

\subsection{Weak Pointers}

A weak pointer provides a way to maintain a reference to an object
without preventing an object from being garbage collected.  If the
garbage collector discovers that the only pointers to an object are
weak pointers, then it breaks the weak pointers and deallocates the
object.

\begin{defun}{extensions:}{make-weak-pointer}{\args{\var{object}}}
  \defunx[extensions:]{weak-pointer-value}{\args{\var{weak-pointer}}}
  
  \code{make-weak-pointer} returns a weak pointer to an object.
  \code{weak-pointer-value} follows a weak pointer, returning the two
  values: the object pointed to (or \false{} if broken) and a boolean
  value which is \false{} if the pointer has been broken, and true
  otherwise.
\end{defun}


\subsection{Finalization}

Finalization provides a ``hook'' that is triggered when the garbage
collector reclaims an object.  It is usually used to recover non-Lisp
resources that were allocated to implement the finalized Lisp object.
For example, when a unix file-descriptor stream is collected,
finalization is used to close the underlying file descriptor.

\begin{defun}{extensions:}{finalize}{\args{\var{object} \var{function}}}
  
  This function registers \var{object} for finalization.
  \var{function} is called with no arguments when \var{object} is
  reclaimed.  Normally \var{function} will be a closure over the
  underlying state that needs to be freed, e.g. the unix file
  descriptor in the fd-stream case.  Note that \var{function} must not
  close over \var{object} itself, as this prevents the object from
  ever becoming garbage.
\end{defun}

\begin{defun}{extensions:}{cancel-finalization}{\args{\var{object}}}
  
  This function cancel any finalization request for \var{object}.
\end{defun}


\section{Describe}

\begin{defun}{}{describe}{ \args{\var{object} \ampoptional{} \var{stream}}}
  
  The \code{describe} function prints useful information about
  \var{object} on \var{stream}, which defaults to
  \code{*standard-output*}.  For any object, \code{describe} will
  print out the type.  Then it prints other information based on the
  type of \var{object}.  The types which are presently handled are:

  \begin{Lentry}
  
  \item[\tindexed{hash-table}] \code{describe} prints the number of
    entries currently in the hash table and the number of buckets
    currently allocated.
  
  \item[\tindexed{function}] \code{describe} prints a list of the
    function's name (if any) and its formal parameters.  If the name
    has function documentation, then it will be printed.  If the
    function is compiled, then the file where it is defined will be
    printed as well.
  
  \item[\tindexed{fixnum}] \code{describe} prints whether the integer
    is prime or not.
  
  \item[\tindexed{symbol}] The symbol's value, properties, and
    documentation are printed.  If the symbol has a function
    definition, then the function is described.
  \end{Lentry}
  If there is anything interesting to be said about some component of
  the object, describe will invoke itself recursively to describe that
  object.  The level of recursion is indicated by indenting output.
\end{defun}

A number of switches can be used to control \code{describe}'s behavior.

\begin{defvar}{extensions:}{describe-level}

  The maximum level of recursive description allowed.  Initially two.
\end{defvar}

\begin{defvar}{extensions:}{describe-indentation}

The number of spaces to indent for each level of recursive
description, initially three.
\end{defvar}

\begin{defvar}{extensions:}{describe-print-level}
  \defvarx[extensions:]{describe-print-length}
  
  The values of \code{*print-level*} and \code{*print-length*} during
  description.  Initially two and five.
\end{defvar}


\section{The Inspector}

\cmucl{} has both a graphical inspector that uses the X Window System,
and a simple terminal-based inspector.

\begin{defun}{}{inspect}{ \args{\ampoptional{} \var{object}}}
  
  \code{inspect} calls the inspector on the optional argument
  \var{object}.  If \var{object} is unsupplied, \code{inspect}
  immediately returns \false.  Otherwise, the behavior of inspect
  depends on whether Lisp is running under X.  When \code{inspect} is
  eventually exited, it returns some selected Lisp object.
\end{defun}


\subsection{The Graphical Interface}
\label{motif-interface}

\cmucl{} has an interface to Motif which is functionally similar to
CLM, but works better in \cmucl{}.  This interface is documented in
separate manuals \textit{CMUCL Motif Toolkit} and \textit{Design Notes
on the Motif Toolkit}, which are distributed with \cmucl{}.

This motif interface has been used to write the inspector and graphical
debugger.  There is also a Lisp control panel with a simple file management
facility, apropos and inspector dialogs, and controls for setting global
options.  See the \code{interface} and \code{toolkit} packages.

\begin{defun}{interface:}{lisp-control-panel}{}
  
  This function creates a control panel for the Lisp process.
\end{defun}

\begin{defvar}{interface:}{interface-style}
  
  When the graphical interface is loaded, this variable controls
  whether it is used by \code{inspect} and the error system.  If the
  value is \kwd{graphics} (the default) and the \code{DISPLAY}
  environment variable is defined, the graphical inspector and
  debugger will be invoked by \findexed{inspect} or when an error is
  signalled.  Possible values are \kwd{graphics} and {tty}.  If the
  value is \kwd{graphics}, but there is no X display, then we quietly
  use the TTY interface.
\end{defvar}


\subsection{The TTY Inspector}

If X is unavailable, a terminal inspector is invoked.  The TTY inspector
is a crude interface to \code{describe} which allows objects to be
traversed and maintains a history.  This inspector prints information
about and object and a numbered list of the components of the object.
The command-line based interface is a normal
\code{read}--\code{eval}--\code{print} loop, but an integer \var{n}
descends into the \var{n}'th component of the current object, and
symbols with these special names are interpreted as commands:

\begin{Lentry}
\item[U] Move back to the enclosing object.  As you descend into the
components of an object, a stack of all the objects previously seen is
kept.  This command pops you up one level of this stack.

\item[Q, E] Return the current object from \code{inspect}.

\item[R] Recompute object display, and print again.  Useful if the
object may have changed.

\item[D] Display again without recomputing.

\item[H, ?] Show help message.
\end{Lentry}


\section{Load}

\begin{defun}{}{load}{%
    \args{\var{filename}
      \keys{\kwd{verbose} \kwd{print} \kwd{if-does-not-exist}}
      \morekeys{\kwd{if-source-newer} \kwd{contents}}}}
  
  As in standard \clisp{}, this function loads a file containing
  source or object code into the running Lisp.  Several CMU extensions
  have been made to \code{load} to conveniently support a variety of
  program file organizations.  \var{filename} may be a wildcard
  pathname such as \file{*.lisp}, in which case all matching files are
  loaded.
  
  If \var{filename} has a \code{pathname-type} (or extension), then
  that exact file is loaded.  If the file has no extension, then this
  tells \code{load} to use a heuristic to load the ``right'' file.
  The \code{*load-source-types*} and \code{*load-object-types*}
  variables below are used to determine the default source and object
  file types.  If only the source or the object file exists (but not
  both), then that file is quietly loaded.  Similarly, if both the
  source and object file exist, and the object file is newer than the
  source file, then the object file is loaded.  The value of the
  \var{if-source-newer} argument is used to determine what action to
  take when both the source and object files exist, but the object
  file is out of date:
  \begin{Lentry}
  \item[\kwd{load-object}] The object file is loaded even though the
    source file is newer.
    
  \item[\kwd{load-source}] The source file is loaded instead of the
    older object file.
    
  \item[\kwd{compile}] The source file is compiled and then the new
    object file is loaded.
    
  \item[\kwd{query}] The user is asked a yes or no question to
    determine whether the source or object file is loaded.
  \end{Lentry}
  This argument defaults to the value of
  \code{ext:*load-if-source-newer*} (initially \kwd{load-object}.)
  
  The \var{contents} argument can be used to override the heuristic
  (based on the file extension) that normally determines whether to
  load the file as a source file or an object file.  If non-null, this
  argument must be either \kwd{source} or \kwd{binary}, which forces
  loading in source and binary mode, respectively. You really
  shouldn't ever need to use this argument.
\end{defun}

\begin{defvar}{extensions:}{load-source-types}
  \defvarx[extensions:]{load-object-types}
  
  These variables are lists of possible \code{pathname-type} values
  for source and object files to be passed to \code{load}.  These
  variables are only used when the file passed to \code{load} has no
  type; in this case, the possible source and object types are used to
  default the type in order to determine the names of the source and
  object files.
\end{defvar}

\begin{defvar}{extensions:}{load-if-source-newer}
  
  This variable determines the default value of the
  \var{if-source-newer} argument to \code{load}.  Its initial value is
  \kwd{load-object}.
\end{defvar}


\section{The Reader}

\subsection{Reader Extensions}
\cmucl{} supports an ANSI-compatible extension to enable reading of
specialized arrays.  Thus
\begin{example}
  * (setf *print-readably* nil)
  NIL
  * (make-array '(2 2) :element-type '(signed-byte 8))
  #2A((0 0) (0 0))
  * (setf *print-readably* t)
  T
  * (make-array '(2 2) :element-type '(signed-byte 8))
  #A((SIGNED-BYTE 8) (2 2) ((0 0) (0 0)))
  * (type-of (read-from-string "#A((SIGNED-BYTE 8) (2 2) ((0 0) (0 0)))"))
  (SIMPLE-ARRAY (SIGNED-BYTE 8) (2 2))
  * (setf *print-readably* nil)
  NIL
  * (type-of (read-from-string "#A((SIGNED-BYTE 8) (2 2) ((0 0) (0 0)))"))
  (SIMPLE-ARRAY (SIGNED-BYTE 8) (2 2))
\end{example}

\subsection{Reader Parameters}
\begin{defvar}{extensions:}{ignore-extra-close-parentheses}
  
  If this variable is \true{} (the default), then the reader merely
  prints a warning when an extra close parenthesis is detected
  (instead of signalling an error.)
\end{defvar}

\section{Stream Extensions}
\begin{defun}{sys:}{read-n-bytes}{%
    \args{\var{stream buffer start numbytes} 
      \ampoptional{} \var{eof-error-p}}}
  
  On streams that support it, this function reads multiple bytes of
  data into a buffer.  The buffer must be a \code{simple-string} or
  \code{(simple-array (unsigned-byte 8) (*))}.  The argument
  \var{nbytes} specifies the desired number of bytes, and the return
  value is the number of bytes actually read.
  \begin{itemize}
  \item If \var{eof-error-p} is true, an \tindexed{end-of-file}
    condition is signalled if end-of-file is encountered before
    \var{count} bytes have been read.
    
  \item If \var{eof-error-p} is false, \code{read-n-bytes reads} as
    much data is currently available (up to count bytes.)  On pipes or
    similar devices, this function returns as soon as any data is
    available, even if the amount read is less than \var{count} and
    eof has not been hit.  See also \funref{make-fd-stream}.
  \end{itemize}
\end{defun}

\input{simple-streams}

\section{Running Programs from Lisp}

It is possible to run programs from Lisp by using the following function.

\begin{defun}{extensions:}{run-program}{%
    \args{\var{program} \var{args}
      \keys{\kwd{env} \kwd{wait} \kwd{pty} \kwd{input}}
      \morekeys{\kwd{if-input-does-not-exist}}
      \yetmorekeys{\kwd{output} \kwd{if-output-exists}}
      \yetmorekeys{\kwd{error} \kwd{if-error-exists}}
      \yetmorekeys{\kwd{status-hook} \kwd{external-format}}
      \yetmorekeys{\kwd{element-type}}}}
     
  \code{run-program} runs \var{program} in a child process.
  \var{Program} should be a pathname or string naming the program.
  \var{Args} should be a list of strings which this passes to
  \var{program} as normal Unix parameters.  For no arguments, specify
  \var{args} as \nil.  The value returned is either a process
  structure or \nil.  The process interface follows the description of
  \code{run-program}.  If \code{run-program} fails to fork the child
  process, it returns \nil.
  
  Except for sharing file descriptors as explained in keyword argument
  descriptions, \code{run-program} closes all file descriptors in the
  child process before running the program.  When you are done using a
  process, call \code{process-close} to reclaim system resources.  You
  only need to do this when you supply \kwd{stream} for one of
  \kwd{input}, \kwd{output}, or \kwd{error}, or you supply \kwd{pty}
  non-\nil.  You can call \code{process-close} regardless of whether
  you must to reclaim resources without penalty if you feel safer.

  \code{run-program} accepts the following keyword arguments:

  \begin{Lentry}   
  \item[\kwd{env}] This is an a-list mapping keywords and
    simple-strings.  The default is \code{ext:*environment-list*}.  If
    \kwd{env} is specified, \code{run-program} uses the value given
    and does not combine the environment passed to Lisp with the one
    specified.
    
  \item[\kwd{wait}] If non-\nil{} (the default), wait until the child
    process terminates.  If \nil, continue running Lisp while the
    child process runs.
    
  \item[\kwd{pty}] This should be one of \true, \nil, or a stream.  If
    specified non-\nil, the subprocess executes under a Unix PTY.
    If specified as a stream, the system collects all output to this
    pty and writes it to this stream.  If specified as \true, the
    \code{process-pty} slot contains a stream from which you can read
    the program's output and to which you can write input for the
    program.  The default is \nil.
    
  \item[\kwd{input}] This specifies how the program gets its input.
    If specified as a string, it is the name of a file that contains
    input for the child process.  \code{run-program} opens the file as
    standard input.  If specified as \nil{} (the default), then
    standard input is the file \file{/dev/null}.  If specified as
    \true, the program uses the current standard input.  This may
    cause some confusion if \kwd{wait} is \nil{} since two processes
    may use the terminal at the same time.  If specified as
    \kwd{stream}, then the \code{process-input} slot contains an
    output stream.  Anything written to this stream goes to the
    program as input.  \kwd{input} may also be an input stream that
    already contains all the input for the process.  In this case
    \code{run-program} reads all the input from this stream before
    returning, so this cannot be used to interact with the process.
    If \kwd{input} is a string stream, it is up to the caller to call
    \code{string-encode} or other function to convert the string to
    the appropriate encoding.  In either case, the least significant 8
    bits of the \code{char-code} of each \code{character} is
    sent to the program.
    
  \item[\kwd{if-input-does-not-exist}] This specifies what to do if
    the input file does not exist.  The following values are valid:
    \nil{} (the default) causes \code{run-program} to return \nil{}
    without doing anything; \kwd{create} creates the named file; and
    \kwd{error} signals an error.
    
  \item[\kwd{output}] This specifies what happens with the program's
    output.  If specified as a pathname, it is the name of a file that
    contains output the program writes to its standard output.  If
    specified as \nil{} (the default), all output goes to
    \file{/dev/null}.  If specified as \true, the program writes to
    the Lisp process's standard output.  This may cause confusion if
    \kwd{wait} is \nil{} since two processes may write to the terminal
    at the same time.  If specified as \kwd{stream}, then the
    \code{process-output} slot contains an input stream from which you
    can read the program's output.  \kwd{output} can also be a stream
    in which case all output from the process is written to this
    stream.  If \kwd{output} is a string-stream, each octet read from
    the program is converted to a character using \code{code-char}.
    It is up to the caller to convert this using the appropriate
    external format to create the desired encoded string.
    
  \item[\kwd{if-output-exists}] This specifies what to do if the
    output file already exists.  The following values are valid:
    \nil{} causes \code{run-program} to return \nil{} without doing
    anything; \kwd{error} (the default) signals an error;
    \kwd{supersede} overwrites the current file; and \kwd{append}
    appends all output to the file.
    
  \item[\kwd{error}] This is similar to \kwd{output}, except the file
    becomes the program's standard error.  Additionally, \kwd{error}
    can be \kwd{output} in which case the program's error output is
    routed to the same place specified for \kwd{output}.  If specified
    as \kwd{stream}, the \code{process-error} contains a stream
    similar to the \code{process-output} slot when specifying the
    \kwd{output} argument.
    
  \item[\kwd{if-error-exists}] This specifies what to do if the error
    output file already exists.  It accepts the same values as
    \kwd{if-output-exists}.
    
  \item[\kwd{status-hook}] This specifies a function to call whenever
    the process changes status.  This is especially useful when
    specifying \kwd{wait} as \nil.  The function takes the process as
    a required argument.

  \item[\kwd{external-format}] This specifies the external format to
    use for streams created for \code{run-program}.  This does not
    apply to string streams passed in as \kwd{input} or \kwd{output}
    parameters.

  \item[\kwd{element-type}] If streams are created \code{run-program},
    use this as the \kwd{element-type} for the stream.  Defaults to
    \code{BASE-CHAR}. 
    
%   \item[\kwd{before-execve}] This specifies a function to run in the
%     child process before it becomes the program to run.  This is
%     useful for actions such as authenticating the child process
%     without modifying the parent Lisp process.
  \end{Lentry}
\end{defun}


\subsection{Process Accessors}

The following functions interface the process returned by \code{run-program}:

\begin{defun}{extensions:}{process-p}{\args{\var{thing}}}
  
  This function returns \true{} if \var{thing} is a process.
  Otherwise it returns \nil{}
\end{defun}

\begin{defun}{extensions:}{process-pid}{\args{\var{process}}}
  
  This function returns the process ID, an integer, for the
  \var{process}.
\end{defun}

\begin{defun}{extensions:}{process-status}{\args{\var{process}}}
  
  This function returns the current status of \var{process}, which is
  one of \kwd{running}, \kwd{stopped}, \kwd{exited}, or
  \kwd{signaled}.
\end{defun}

\begin{defun}{extensions:}{process-exit-code}{\args{\var{process}}}
  
  This function returns either the exit code for \var{process}, if it
  is \kwd{exited}, or the termination signal \var{process} if it is
  \kwd{signaled}.  The result is undefined for processes that are
  still alive.
\end{defun}

\begin{defun}{extensions:}{process-core-dumped}{\args{\var{process}}}
  
  This function returns \true{} if someone used a Unix signal to
  terminate the \var{process} and caused it to dump a Unix core image.
\end{defun}

\begin{defun}{extensions:}{process-pty}{\args{\var{process}}}
  
  This function returns either the two-way stream connected to
  \var{process}'s Unix PTY connection or \nil{} if there is none.
\end{defun}

\begin{defun}{extensions:}{process-input}{\args{\var{process}}}
  \defunx[extensions:]{process-output}{\args{\var{process}}}
  \defunx[extensions:]{process-error}{\args{\var{process}}}
  
  If the corresponding stream was created, these functions return the
  input, output or error fd-stream.  \nil{} is returned if there
  is no stream.
\end{defun}

\begin{defun}{extensions:}{process-status-hook}{\args{\var{process}}}
  
  This function returns the current function to call whenever
  \var{process}'s status changes.  This function takes the
  \var{process} as a required argument.  \code{process-status-hook} is
  \code{setf}'able.
\end{defun}

\begin{defun}{extensions:}{process-plist}{\args{\var{process}}}
  
  This function returns annotations supplied by users, and it is
  \code{setf}'able.  This is available solely for users to associate
  information with \var{process} without having to build a-lists or
  hash tables of process structures.
\end{defun}

\begin{defun}{extensions:}{process-wait}{
    \args{\var{process} \ampoptional{} \var{check-for-stopped}}}
  
  This function waits for \var{process} to finish.  If
  \var{check-for-stopped} is non-\nil, this also returns when
  \var{process} stops.
\end{defun}

\begin{defun}{extensions:}{process-kill}{%
    \args{\var{process} \var{signal} \ampoptional{} \var{whom}}}
  
  This function sends the Unix \var{signal} to \var{process}.
  \var{Signal} should be the number of the signal or a keyword with
  the Unix name (for example, \kwd{sigsegv}).  \var{Whom} should be
  one of the following:
  \begin{Lentry}
    
  \item[\kwd{pid}] This is the default, and it indicates sending the
    signal to \var{process} only.
    
  \item[\kwd{process-group}] This indicates sending the signal to
    \var{process}'s group.
    
  \item[\kwd{pty-process-group}] This indicates sending the signal to
    the process group currently in the foreground on the Unix PTY
    connected to \var{process}.  This last option is useful if the
    running program is a shell, and you wish to signal the program
    running under the shell, not the shell itself.  If
    \code{process-pty} of \var{process} is \nil, using this option is
    an error.
  \end{Lentry}
\end{defun}

\begin{defun}{extensions:}{process-alive-p}{\args{\var{process}}}
  
  This function returns \true{} if \var{process}'s status is either
  \kwd{running} or \kwd{stopped}.
\end{defun}

\begin{defun}{extensions:}{process-close}{\args{\var{process}}}
  
  This function closes all the streams associated with \var{process}.
  When you are done using a process, call this to reclaim system
  resources.
\end{defun}


\section{Saving a Core Image}

A mechanism has been provided to save a running Lisp core image and to
later restore it.  This is convenient if you don't want to load several files
into a Lisp when you first start it up.  The main problem is the large
size of each saved Lisp image, typically at least 20 megabytes.

\begin{defun}{extensions:}{save-lisp}{%
    \args{\var{file}
      \keys{\kwd{purify} \kwd{root-structures} \kwd{init-function}}
      \morekeys{\kwd{load-init-file} \kwd{print-herald} \kwd{site-init}}
      \yetmorekeys{\kwd{process-command-line} \kwd{batch-mode} \kwd{executable}}}}
  
  The \code{save-lisp} function saves the state of the currently
  running Lisp core image in \var{file}.  The keyword arguments have
  the following meaning:
  \begin{Lentry}
    
  \item[\kwd{purify}] If non-\nil{} (the default), the core image is
    purified before it is saved (see \funref{purify}.)  This reduces
    the amount of work the garbage collector must do when the
    resulting core image is being run.  Also, if more than one Lisp is
    running on the same machine, this maximizes the amount of memory
    that can be shared between the two processes.
    
  \item[\kwd{root-structures}]
      This should be a list of the main entry points in any newly
      loaded systems.  This need not be supplied, but locality and/or
      GC performance will be better if they are.  Meaningless if
      \kwd{purify} is \nil.  See \funref{purify}.

  \item[\kwd{init-function}] This is the function that starts running
    when the created core file is resumed.  The default function
    simply invokes the top level read-eval-print loop.  If the
    function returns the lisp will exit.
    
  \item[\kwd{load-init-file}] If non-NIL, then load an init file;
    either the one specified on the command line or
    ``\w{\file{init.}\var{fasl-type}}'', or, if
    ``\w{\file{init.}\var{fasl-type}}'' does not exist,
    \code{init.lisp} from the user's home directory.  If the init file
    is found, it is loaded into the resumed core file before the
    read-eval-print loop is entered.
    
  \item[\kwd{site-init}] If non-NIL, the name of the site init file to
    quietly load.  The default is \file{library:site-init}.  No error
    is signalled if the file does not exist.
    
  \item[\kwd{print-herald}] If non-NIL (the default), then print out
    the standard Lisp herald when starting.
    
  \item[\kwd{process-command-line}] If non-NIL (the default),
    processes the command line switches and performs the appropriate
    actions.

  \item[\kwd{batch-mode}] If NIL (the default), then the presence of
    the -batch command-line switch will invoke batch-mode processing
    upon resuming the saved core.  If non-NIL, the produced core will
    always be in batch-mode, regardless of any command-line switches.

  \item[\kwd{executable}] If non-NIL, an executable image is created.
    Normally, \cmucl{} consists of the C runtime along with a core
    file image.  When \kwd{executable} is non-NIL, the core file is
    incorporated into the C runtime, so one (large) executable is
    created instead of a new separate core file.

    This feature is only available on some platforms, as indicated by
    having the feature \kwd{executable}.  Currently only x86 ports and
    the solaris/sparc port have this feature.
  \end{Lentry}
\end{defun}

To resume a saved file, type:
\begin{example}
lisp -core file
\end{example}
However, if the \kwd{executable} option was specified, you can just
use
\begin{example}
  file
\end{example}
since the executable contains the core file within the executable.

\begin{defun}{extensions:}{purify}{
    \args{\var{file}
      \keys{\kwd{root-structures} \kwd{environment-name}}}}
  
  This function optimizes garbage collection by moving all currently
  live objects into non-collected storage.  Once statically allocated,
  the objects can never be reclaimed, even if all pointers to them are
  dropped.  This function should generally be called after a large
  system has been loaded and initialized.

  \begin{Lentry}
  \item[\kwd{root-structures}] is an optional list of objects which
    should be copied first to maximize locality.  This should be a
    list of the main entry points for the resulting core image.  The
    purification process tries to localize symbols, functions, etc.,
    in the core image so that paging performance is improved.  The
    default value is NIL which means that Lisp objects will still be
    localized but probably not as optimally as they could be.
  
    \var{defstruct} structures defined with the \code{(:pure t)}
    option are moved into read-only storage, further reducing GC cost.
    List and vector slots of pure structures are also moved into
    read-only storage.
  
  \item[\kwd{environment-name}] is gratuitous documentation for the
    compacted version of the current global environment (as seen in
    \code{c::*info-environment*}.)  If \false{} is supplied, then
    environment compaction is inhibited.
  \end{Lentry}
\end{defun}


\section{Pathnames}

In \clisp{} quite a few aspects of \tindexed{pathname} semantics are left to
the implementation.  


\subsection{Unix Pathnames}
\cpsubindex{unix}{pathnames}

Unix pathnames are always parsed with a \code{unix-host} object as the host and
\code{nil} as the device.  The last two dots (\code{.}) in the namestring mark
the type and version, however if the first character is a dot, it is considered
part of the name.  If the last character is a dot, then the pathname has the
empty-string as its type.  The type defaults to \code{nil} and the version
defaults to \kwd{newest}.

\begin{example}
(defun parse (x)
  (values (pathname-name x) (pathname-type x) (pathname-version x)))

(parse "foo") \result "foo", NIL, NIL
(parse "foo.bar") \result "foo", "bar", NIL
(parse ".foo") \result ".foo", NIL, NIL
(parse ".foo.bar") \result ".foo", "bar", NIL
(parse "..") \result NIL, NIL, NIL
(parse "foo.") \result "foo", "", NIL
(parse "foo.bar.~1~") \result "foo", "bar", 1
(parse "foo.bar.baz") \result "foo.bar", "baz", NIL
\end{example}

The directory of pathnames beginning with a slash (or a search-list,
\pxlref{search-lists}) is starts \kwd{absolute}, others start with
\kwd{relative}.  The \code{..} directory is parsed as \kwd{up}; there is no
namestring for \kwd{back}:

\begin{example}
(pathname-directory "/usr/foo/bar.baz") \result (:ABSOLUTE "usr" "foo")
(pathname-directory "../foo/bar.baz") \result (:RELATIVE :UP "foo")
\end{example}


\subsection{Wildcard Pathnames}

Wildcards are supported in Unix pathnames.  If `\code{*}' is specified for a
part of a pathname, that is parsed as \kwd{wild}.  `\code{**}' can be used as a
directory name to indicate \kwd{wild-inferiors}.  Filesystem operations
treat \kwd{wild-inferiors} the same as\ \kwd{wild}, but pathname pattern
matching (e.g. for logical pathname translation, \pxlref{logical-pathnames})
matches any number of directory parts with `\code{**}' (see
\pxlref{wildcard-matching}.)

`\code{*}' embedded in a pathname part matches any number of characters.
Similarly, `\code{?}' matches exactly one character, and `\code{[a,b]}'
matches the characters `\code{a}' or `\code{b}'.  These pathname parts are
parsed as \code{pattern} objects.

Backslash can be used as an escape character in namestring
parsing to prevent the next character from being treated as a wildcard.  Note
that if typed in a string constant, the backslash must be doubled, since the
string reader also uses backslash as a quote:

\begin{example}
(pathname-name "foo\(\backslash\backslash\)*bar") => "foo*bar"
\end{example}


\subsection{Logical Pathnames}
\cindex{logical pathnames}
\label{logical-pathnames}

If a namestring begins with the name of a defined logical pathname
host followed by a colon, then it will be parsed as a logical
pathname.  Both `\code{*}' and `\code{**}' wildcards are implemented.
\findexed{load-logical-pathname-translations} on \var{name} looks for a
logical host definition file in
\w{\file{library:\var{name}.translations}}. Note that \file{library:}
designates the search list (\pxlref{search-lists}) initialized to the
\cmucl{} \file{lib/} directory, not a logical pathname.  The format of
the file is a single list of two-lists of the from and to patterns:

\begin{example}
(("foo;*.text" "/usr/ram/foo/*.txt")
 ("foo;*.lisp" "/usr/ram/foo/*.l"))
\end{example}


\subsection{Search Lists}
\cindex{search lists}
\label{search-lists}

Search lists are an extension to \clisp{} pathnames.  They serve a function
somewhat similar to \clisp{} logical pathnames, but work more like Unix PATH
variables.  Search lists are used for two purposes:
\begin{itemize}
\item They provide a convenient shorthand for commonly used directory names,
and

\item They allow the abstract (directory structure independent) specification
of file locations in program pathname constants (similar to logical pathnames.)
\end{itemize}
Each search list has an associated list of directories (represented as
pathnames with no name or type component.)  The namestring for any relative
pathname may be prefixed with ``\var{slist}\code{:}'', indicating that the
pathname is relative to the search list \var{slist} (instead of to the current
working directory.)  Once qualified with a search list, the pathname is no
longer considered to be relative.

When a search list qualified pathname is passed to a file-system operation such
as \code{open}, \code{load} or \code{truename}, each directory in the search
list is successively used as the root of the pathname until the file is
located.  When a file is written to a search list directory, the file is always
written to the first directory in the list.


\subsection{Predefined Search-Lists}

These search-lists are initialized from the Unix environment or when Lisp was
built:
\begin{Lentry}
\item[\code{default:}] The current directory at startup.

\item[\code{home:}] The user's home directory.

\item[\code{library:}] The \cmucl{} \file{lib/} directory (\code{CMUCLLIB} environment
variable).

\item[\code{path:}] The Unix command path (\code{PATH} environment variable).
\item[\code{ld-library-path:}] The Unix \code{LD\_LIBRARY\_PATH}
  environment variable.
\item[\code{target:}] The root of the tree where \cmucl{} was compiled.
\item[\code{modules:}] The list of directories where \cmucl{}'s
  modules can be found.
\item[\code{ext-formats:}] The list of directories where \cmucl{} can
  find the implementation of external formats.  
\end{Lentry}
It can be useful to redefine these search-lists, for example, \file{library:}
can be augmented to allow logical pathname translations to be located, and
\file{target:} can be redefined to point to where \cmucl{} system sources are
locally installed. 


\subsection{Search-List Operations}

These operations define and access search-list definitions.  A search-list name
may be parsed into a pathname before the search-list is actually defined, but
the search-list must be defined before it can actually be used in a filesystem
operation.

\begin{defun}{extensions:}{search-list}{\var{name}}
  
  This function returns the list of directories associated with the
  search list \var{name}.  If \var{name} is not a defined search list,
  then an error is signaled.  When set with \code{setf}, the list of
  directories is changed to the new value.  If the new value is just a
  namestring or pathname, then it is interpreted as a one-element
  list.  Note that (unlike Unix pathnames), search list names are
  case-insensitive.
\end{defun}

\begin{defun}{extensions:}{search-list-defined-p}{\var{name}}
  \defunx[extensions:]{clear-search-list}{\var{name}}
  
  \code{search-list-defined-p} returns \true{} if \var{name} is a
  defined search list name, \false{} otherwise.
  \code{clear-search-list} make the search list \var{name} undefined.
\end{defun}

\begin{defmac}{extensions:}{enumerate-search-list}{%
    \args{(\var{var} \var{pathname} \mopt{result}) \mstar{form}}}
  
  This macro provides an interface to search list resolution.  The
  body \var{forms} are executed with \var{var} bound to each
  successive possible expansion for \var{name}.  If \var{name} does
  not contain a search-list, then the body is executed exactly once.
  Everything is wrapped in a block named \nil, so \code{return} can be
  used to terminate early.  The \var{result} form (default \nil) is
  evaluated to determine the result of the iteration.
\end{defmac}


\subsection{Search List Example}

The search list \code{code:} can be defined as follows:
\begin{example}
(setf (ext:search-list "code:") '("/usr/lisp/code/"))
\end{example}
It is now possible to use \code{code:} as an abbreviation for the directory
\file{/usr/lisp/code/} in all file operations.  For example, you can now specify
\code{code:eval.lisp} to refer to the file \file{/usr/lisp/code/eval.lisp}.

To obtain the value of a search-list name, use the function search-list
as follows:
\begin{example}
(ext:search-list \var{name})
\end{example}
Where \var{name} is the name of a search list as described above.  For example,
calling \code{ext:search-list} on \code{code:} as follows:
\begin{example}
(ext:search-list "code:")
\end{example}
returns the list \code{("/usr/lisp/code/")}.


\section{Filesystem Operations}

\cmucl{} provides a number of extensions and optional features beyond those
required by the \clisp{} specification.


\subsection{Wildcard Matching}
\label{wildcard-matching}

Unix filesystem operations such as \code{open} will accept wildcard pathnames
that match a single file (of course, \code{directory} allows any number of
matches.)  Filesystem operations treat \kwd{wild-inferiors} the same as\
\kwd{wild}.

\begin{defun}{}{directory}{\var{wildname} \keys{\kwd{all} \kwd{check-for-subdirs}}
    \kwd{truenamep} \morekeys{\kwd{follow-links}}}
  
  The keyword arguments to this \clisp{} function are a \cmucl{} extension.
  The arguments (all default to \code{t}) have the following
  functions:
  \begin{Lentry}
  \item[\kwd{all}] Include files beginning with dot such as
    \file{.login}, similar to ``\code{ls -a}''.
    
  \item[\kwd{check-for-subdirs}] Test whether files are directories,
    similar to ``\code{ls -F}''.
    
  \item[\kwd{truenamep}] Call \code{truename} on each file, which
    expands out all symbolic links.  Note that this option can easily
    result in pathnames being returned which have a different
    directory from the one in the \var{wildname} argument.

  \item[\kwd{follow-links}] Follow symbolic links when searching for
    matching directories.
  \end{Lentry}
\end{defun}

\begin{defun}{extensions:}{print-directory}{%
    \args{\var{wildname}
      \ampoptional{} \var{stream}
      \keys{\kwd{all} \kwd{verbose}}
      \morekeys{\kwd{return-list}}}}
  
  Print a directory of \var{wildname} listing to \var{stream} (default
  \code{*standard-output*}.)  \kwd{all} and \kwd{verbose} both default
  to \false{} and correspond to the ``\code{-a}'' and ``\code{-l}''
  options of \file{ls}.  Normally this function returns \false{}, but
  if \kwd{return-list} is true, a list of the matched pathnames are
  returned.
\end{defun}


\subsection{File Name Completion}

\begin{defun}{extensions:}{complete-file}{%
    \args{\var{pathname}
      \keys{\kwd{defaults} \kwd{ignore-types}}}}
  
  Attempt to complete a file name to the longest unambiguous prefix.
  If supplied, directory from \kwd{defaults} is used as the ``working
  directory'' when doing completion.  \kwd{ignore-types} is a list of
  strings of the pathname types (a.k.a. extensions) that should be
  disregarded as possible matches (binary file names, etc.)
\end{defun}

\begin{defun}{extensions:}{ambiguous-files}{%
    \args{\var{pathname}
      \ampoptional{} \var{defaults}}}
  
  Return a list of pathnames for all the possible completions of
  \var{pathname} with respect to \var{defaults}.
\end{defun}


\subsection{Miscellaneous Filesystem Operations}

\begin{defun}{extensions:}{default-directory}{}
  
  Return the current working directory as a pathname.  If set with
  \code{setf}, set the working directory.
\end{defun}

\begin{defun}{extensions:}{file-writable}{\var{name}}
  
  This function accepts a pathname and returns \true{} if the current
  process can write it, and \false{} otherwise.
\end{defun}

\begin{defun}{extensions:}{unix-namestring}{%
    \args{\var{pathname}
      \ampoptional{} \var{for-input}}}
  
  This function converts \var{pathname} into a string that can be used
  with UNIX system calls.  Search-lists and wildcards are expanded.
  \var{for-input} controls the treatment of search-lists: when true
  (the default) and the file exists anywhere on the search-list, then
  that absolute pathname is returned; otherwise the first element of
  the search-list is used as the directory.
\end{defun}


\section{Time Parsing and Formatting}

\cindex{time parsing} \cindex{time formatting}
Functions are provided to allow parsing strings containing time information
and printing time in various formats are available.

\begin{defun}{extensions:}{parse-time}{%
    \args{\var{time-string}
      \keys{\kwd{error-on-mismatch} \kwd{default-seconds}}
      \morekeys{\kwd{default-minutes} \kwd{default-hours}}
      \yetmorekeys{\kwd{default-day} \kwd{default-month}}
      \yetmorekeys{\kwd{default-year} \kwd{default-zone}}
      \yetmorekeys{\kwd{default-weekday}}}}
  
  \code{parse-time} accepts a string containing a time (e.g.,
  \w{"\code{Jan 12, 1952}"}) and returns the universal time if it is
  successful.  If it is unsuccessful and the keyword argument
  \kwd{error-on-mismatch} is non-\nil{}, it signals an error.
  Otherwise it returns \nil{}.  The other keyword arguments have the
  following meaning:

  \begin{Lentry}
  \item[\kwd{default-seconds}] specifies the default value for the
    seconds value if one is not provided by \var{time-string}.  The
    default value is 0.
    
  \item[\kwd{default-minutes}] specifies the default value for the
    minutes value if one is not provided by \var{time-string}.  The
    default value is 0.
    
  \item[\kwd{default-hours}] specifies the default value for the hours
    value if one is not provided by \var{time-string}.  The default
    value is 0.
    
  \item[\kwd{default-day}] specifies the default value for the day
    value if one is not provided by \var{time-string}.  The default
    value is the current day.
    
  \item[\kwd{default-month}] specifies the default value for the month
    value if one is not provided by \var{time-string}.  The default
    value is the current month.
    
  \item[\kwd{default-year}] specifies the default value for the year
    value if one is not provided by \var{time-string}.  The default
    value is the current year.
    
  \item[\kwd{default-zone}] specifies the default value for the time
    zone value if one is not provided by \var{time-string}.  The
    default value is the current time zone.
    
  \item[\kwd{default-weekday}] specifies the default value for the day
    of the week if one is not provided by \var{time-string}.  The
    default value is the current day of the week.
  \end{Lentry}
  Any of the above keywords can be given the value \kwd{current} which
  means to use the current value as determined by a call to the
  operating system.
\end{defun}

\begin{defun}{extensions:}{format-universal-time}{
    \args{\var{dest} \var{universal-time}
       \\
       \keys{\kwd{timezone}}
       \morekeys{\kwd{style} \kwd{date-first}}
       \yetmorekeys{\kwd{print-seconds} \kwd{print-meridian}}
       \yetmorekeys{\kwd{print-timezone} \kwd{print-weekday}}}}
   \defunx[extensions:]{format-decoded-time}{
     \args{\var{dest} \var{seconds} \var{minutes} \var{hours} \var{day} \var{month} \var{year}
       \\
       \keys{\kwd{timezone}}
       \morekeys{\kwd{style} \kwd{date-first}}
       \yetmorekeys{\kwd{print-seconds} \kwd{print-meridian}}
       \yetmorekeys{\kwd{print-timezone} \kwd{print-weekday}}}}
   
   \code{format-universal-time} formats the time specified by
   \var{universal-time}.  \code{format-decoded-time} formats the time
   specified by \var{seconds}, \var{minutes}, \var{hours}, \var{day},
   \var{month}, and \var{year}.  \var{Dest} is any destination
   accepted by the \code{format} function.  The keyword arguments have
   the following meaning:
   \begin{Lentry}
     
   \item[\kwd{timezone}] is an integer specifying the hours west of
     Greenwich.  \kwd{timezone} defaults to the current time zone.
     
   \item[\kwd{style}] specifies the style to use in formatting the
     time.  The legal values are:
     \begin{Lentry}
  
     \item[\kwd{short}] specifies to use a numeric date.
  
     \item[\kwd{long}] specifies to format months and weekdays as
       words instead of numbers.
  
     \item[\kwd{abbreviated}] is similar to long except the words are
       abbreviated.
  
     \item[\kwd{government}] is similar to abbreviated, except the
       date is of the form ``day month year'' instead of ``month day,
       year''.
     \end{Lentry}
     
   \item[\kwd{date-first}] if non-\false{} (default) will place the
     date first.  Otherwise, the time is placed first.
  
   \item[\kwd{print-seconds}] if non-\false{} (default) will format
     the seconds as part of the time.  Otherwise, the seconds will be
     omitted.
  
   \item[\kwd{print-meridian}] if non-\false{} (default) will format
     ``AM'' or ``PM'' as part of the time.  Otherwise, the ``AM'' or
     ``PM'' will be omitted.
  
   \item[\kwd{print-timezone}] if non-\false{} (default) will format
     the time zone as part of the time.  Otherwise, the time zone will
     be omitted.

     %%\item[\kwd{print-seconds}]
     %%if non-\false{} (default) will format the seconds as part of
     %%the time.  Otherwise, the seconds will be omitted.
  
   \item[\kwd{print-weekday}] if non-\false{} (default) will format
     the weekday as part of date.  Otherwise, the weekday will be
     omitted.
   \end{Lentry}
\end{defun}


\section{Random Number Generation}
\cindex{random number generation}

\clisp{} includes a random number generator as a standard part of the
language; however, the implementation of the generator is not
specified.

\subsection{MT-19937 Generator}
\cpsubindex{random number generation}{MT-19937 generator}
On all platforms, the random number is \code{MT-19937} generator as indicated by
\kwd{rand-mt19937} being in \code{*features*}.  This is a Lisp
implementation of the MT-19937 generator of Makoto Matsumoto and
T. Nishimura.  We refer the reader to their paper\footnote{``Mersenne
  Twister: A 623-Dimensionally Equidistributed Uniform Pseudorandom
  Number Generator,'' ACM Trans. on Modeling and Computer Simulation,
  Vol. 8, No. 1, January 1998, pp.3--30} or to
their
\ifpdf
\href{http://www.math.sci.hiroshima-u.ac.jp/~m-mat/MT/emt.html}{website}.
\else
website at
\href{http://www.math.keio.ac.jp/~matumoto/emt.html}{\texttt{http://www.math.keio.ac.jp/~matsumoto/emt.html}}.
\fi

When \cmucl{} starts up, \code{*random-state*} is initialized by
reading 627 words from \code{/dev/urandom}, when available.  If
\code{/dev/urandom} is not available, the universal time is used to
initialize \code{*random-state*}.  The initialization is done as given
in Matsumoto's paper.

\section{Lisp Threads}
\cindex{lisp threads}

\cmucl{} supports Lisp threads for the x86 platform.

\section{Lisp Library}
\label{lisp-lib}

The \cmucl{} project maintains a collection of useful or interesting
programs written by users of our system.  The library is in
\file{lib/contrib/}.  Two files there that users should read are:
\begin{Lentry}

\item[CATALOG.TXT]
This file contains a page for each entry in the library.  It
contains information such as the author, portability or dependency issues, how
to load the entry, etc.

\item[READ-ME.TXT]
This file describes the library's organization and all the
possible pieces of information an entry's catalog description could contain.
\end{Lentry}

Hemlock has a command \F{Library Entry} that displays a list of the current
library entries in an editor buffer.  There are mode specific commands that
display catalog descriptions and load entries.  This is a simple and convenient
way to browse the library.


\section{Generalized Function Names}

\begin{defmac}{ext:}{define-function-name-syntax}{%
    \var{name} (\var{var}) \ampbody\ \var{body}}
  Define lists starting with the symbol \code{name} as a new extended
  function name syntax.
  
  \code{body} is executed with \code{var} bound to an actual function
  name of that form, and should return two values:

  \begin{itemize}
  \item A generalized boolean that is true if \code{var} is a valid
    function name.
  \item A symbol that can be used as a \code{block} name in functions
    whose name is \code{var}.  (For some sorts of function names it
    might make sense to return \code{nil} for the block name, or just
    return one value.)
  \end{itemize}
  
  Users should not define function names starting with a symbol that
  \cmucl{} might be using internally.  It is therefore advisable to
  only define new function names starting with a symbol from a
  user-defined package.
\end{defmac}

\begin{defun}{ext:}{valid-function-name-p}{\var{name}}
  Returns two values:

  \begin{itemize}
  \item True if \code{name} is a valid function name.
  \item A symbol that can be used as a \code{block} name in
    functions whose name is \code{name}.  This can be \code{nil}
    for some function names.
  \end{itemize}
\end{defun}



\section{CLOS}

\subsection{Primary Method Errors}
\cindex{primary method}

The standard requires that an error is signaled when a generic
function is called and

\begin{itemize}
\item no primary method is applicable to the generic function's actual
  arguments, and
\item the generic function's method combination is either the standard
  method combination or a method combination defined with the short
  form of \code{define-method-combination}.  The latter includes the
  standardized method combinations like \code{progn}, \code{and}, etc.
\end{itemize}

\begin{defgeneric}[-generic]{pcl:}{no-primary-method}{\var{gf} \amprest{} \var{args}}
  In \cmucl, this generic function is called in the above erroneous
  cases.  The parameter \code{gf} is the generic function being
  called, and \code{args} is a list of actual arguments in the generic
  function call.
\end{defgeneric}

\begin{defmethod}[-standard]{pcl:}{no-primary-method}{%
    (\var{gf} \argtype{standard-generic-function}) \amprest{} \var{args}}
  This method signals a continuable error of type
  \code{pcl:no-primary-method-error}.
\end{defmethod}


\subsection{Slot Type Checking}
\cindex{slot type checking}

Declared slot types are used when 

\begin{itemize}
\item reading slot values with \code{slot-value} in methods, or

\item setting slots with \code{(setf slot-value)} in methods, or 
  
\item creating instances with \code{make-instance}, when slots are
  initialized from initforms.  This currently depends on PCL being
  able to use its internal \code{make-instance} optimization, which it
  usually can.
\end{itemize}

Example:

\begin{example}
(defclass foo ()
  ((a :type fixnum)))

(defmethod bar ((object foo) value)
  (with-slots (a) object
    (setf a value)))

(defmethod baz ((object foo))
  (< (slot-value object 'a) 10))
\end{example}

In method \code{bar}, and with a suitable safety setting, a type error
will occur if \code{value} is not a \code{fixnum}.  In method
\code{baz}, a \code{fixnum} comparison can be used by the compiler.
  
\begin{defvar}{pcl::}{use-slot-types-p}
  Slot type checking can be turned off by setting this variable to
  \false, which can be useful for compiling code containing incorrect
  slot type declarations.
\end{defvar}


\subsection{Slot Access Optimization}
\cindex{slot access optimization}
\cindex{slot declarations}

The declaration \code{ext:slots} is used for optimizing slot access in
methods.

\begin{example}
declare (ext:slots specifier*)

specifier   ::= (quality class-entry*)
quality     ::= SLOT-BOUNDP | INLINE
class-entry ::= class | (class slot-name*)
class       ::= the name of a class
slot-name   ::= the name of a slot
\end{example}

The \code{slot-boundp} quality specifies that all or some slots of a
class are always bound.

The \code{inline} quality specifies that access to all or some slots
of a class should be inlined, using compile-time knowledge of class
layouts.



\subsubsection{\code{slot-boundp} Declaration}
\cpsubindex{slot declaration}{slot-boundp}

Example:

\begin{example}
(defclass foo ()
  (a b))

(defmethod bar ((x foo))
  (declare (ext:slots (slot-boundp foo)))
  (list (slot-value x 'a) (slot-value x 'b)))
\end{example}

The \code{slot-boundp} declaration in method \code{bar} specifies that
the slots \code{a} and \code{b} accessed through parameter \code{x} in
the scope of the declaration are always bound, because parameter
\code{x} is specialized on class \code{foo} to which the
\code{slot-boundp} declaration applies.  The PCL-generated code for
the \code{slot-value} forms will thus not contain tests for the slots
being bound or not.  The consequences are undefined should one of the
accessed slots not be bound.



\subsubsection{\code{inline} Declaration}
\cpsubindex{slot declaration}{inline}

Example:

\begin{example}
(defclass foo ()
  (a b))

(defmethod bar ((x foo))
  (declare (ext:slots (inline (foo a))))
  (list (slot-value x 'a) (slot-value x 'b)))
\end{example}

The \code{inline} declaration in method \code{bar} tells PCL to use
compile-time knowledge of slot locations for accessing slot \code{a}
of class \code{foo}, in the scope of the declaration.

Class \code{foo} must be known at compile time for this optimization
to be possible.  PCL prints a warning and uses normal slot access If
the class is not defined at compile time.

If a class is \code{proclaim}ed to use inline slot access before it is
defined, the class is defined at compile time.  Example:

\begin{example}
(declaim (ext:slots (inline (foo slot-a))))
(defclass foo () ...)
(defclass bar (foo) ...)
\end{example}
  
Class \code{foo} will be defined at compile time because it is
declared to use inline slot access; methods accessing slot
\code{slot-a} of \code{foo} will use inline slot access if otherwise
possible.  Class \code{bar} will be defined at compile time because
its superclass \code{foo} is declared to use inline slot access.  PCL
uses compile-time information from subclasses to warn about situations
where using inline slot access is not possible.

Normal slot access will be used if PCL finds, at method compilation
time, that

\begin{itemize}
\item class \code{foo} has a subclass in which slot \code{a} is at a
  different location, or

\item there exists a \code{slot-value-using-class} method for
  \code{foo} or a subclass of \code{foo}.
\end{itemize}
  
When the declaration is used to optimize calls to slot accessor
generic functions in methods, as opposed to \code{slot-value} or
\code{(setf slot-value)}, the optimization is additionally not used if

\begin{itemize}
\item there exist, at compile time, applicable methods on the
  reader/writer generic function that are not standard accessor
  methods (for instance, there exist around-methods), or
  
\item applicable reader/writer methods access different slots in a
  class accessed inline, and one of its subclasses.
\end{itemize}

The consequences are undefined if the compile-time environment is not
the same as the run-time environment in these respects, or if the
definition of class \code{foo} or any subclass of \code{foo} is
changed in an incompatible way, that is, if slot locations change.

The effect of the \code{inline} optimization combined with the
\code{slot-boundp} optimization is that CLOS slot access becomes as
fast as structure slot access, which is an order of magnitude faster
than normal CLOS slot access.

\begin{defvar}{pcl::}{optimize-inline-slot-access-p}
  This variable controls if inline slot access optimizations are
  performed.  It is true by default.
\end{defvar}



\subsubsection{Automatic Method Recompilation}
\cindex{methods}
\cpsubindex{methods}{auto-compilation}
\cpsubindex{slot declaration}{method recompilation}
  
Methods using inline slot access can be automatically recompiled after
class changes.  Two declarations control which methods are
automatically recompiled.

\begin{example}
declaim (ext:auto-compile specifier*)
declaim (ext:not-auto-compile specifier*)

specifier   ::= gf-name | (gf-name qualifier* (specializer*))
gf-name     ::= the name of a generic function
qualifier   ::= a method qualifier
specializer ::= a method specializer
\end{example}

If no specifier is given, auto-compilation is by default done/not done
for all methods of all generic functions using inline slot access;
current default is that it is not done.  This global policy can be
overridden on a generic function and method basis.  If
\code{specifier} is a generic function name, it applies to all methods
of that generic function.

Examples:

\begin{example}
(declaim (ext:auto-compile foo))
(defmethod foo :around ((x bar)) ...)
\end{example}

The around-method \code{foo} will be automatically recompiled because
the declamation applies to all methods with name \code{foo}.

\begin{example}
(declaim (ext:auto-compile (foo (bar))))
(defmethod foo :around ((x bar)) ...)
(defmethod foo ((x bar)) ...)
\end{example}

The around-method will not be automatically recompiled, but the
primary method will.

\begin{example}
(declaim (ext:auto-compile foo))
(declaim (ext:not-auto-compile (foo :around (bar)))  
(defmethod foo :around ((x bar)) ...)
(defmethod foo ((x bar)) ...)
\end{example}

The around-method will not be automatically recompiled, because it
is explicitly declaimed not to be.  The primary method will be
automatically recompiled because the first declamation applies to
it.

Auto-recompilation works by recording method bodies using inline slot
access.  When PCL determines that a recompilation is necessary, a
\code{defmethod} form is constructed and evaluated.

Auto-compilation can only be done for methods defined in a null
lexical environment.  PCL prints a warning and doesn't record the
method body if a method using inline slot access is defined in a
non-null lexical environment.  Instead of doing a recompilation on
itself, PCL will then print a warning that the method must be
recompiled manually when classes are changed.



\subsection{Inlining Methods in Effective Methods}
\cindex{effective method}
\cpsubindex{methods}{inlining in effective methods}
\cpsubindex{effective method}{inlining of methods}
\cindex{inline}

When a generic function is called, an effective method is constructed
from applicable methods.  The effective method is called with the
original arguments, and itself calls applicable methods according to
the generic function's method combination.  Some of the function call
overhead in effective methods can be removed by inlining methods in
effective methods, at the expense of increased code size.

Inlining of methods is controlled by the usual \code{inline}
declaration.  In the following example, both \code{foo} methods shown
will be inlined in effective methods:

\begin{example}
(declaim (inline (method foo (foo))
                 (method foo :before (foo))))
(defmethod foo ((x foo)) ...)
(defmethod foo :before ((x foo)) ...)
\end{example}

Please note that this form of inlining has no noticeable effect for
effective methods that consist of a primary method only, which doesn't
have keyword arguments.  In such cases, PCL uses the primary method
directly for the effective method.

When the definition of an inlined method is changed, effective methods
are \textbf{not} automatically updated to reflect the change.  This is
just as it is when inlining normal functions.  Different from the
normal case is that users do not have direct access to effective
methods, as it would be the case when a function is inlined somewhere
else.  Because of this, the function \code{pcl:flush-emf-cache} is
provided for forcing such an update of effective methods.

\begin{defun}{pcl:}{flush-emf-cache}{\ampoptional{} \var{gf}}
  Flush cached effective method functions.  If \code{gf} is supplied,
  it should be a generic function metaobject or the name of a generic
  function, and this function flushes all cached effective methods for
  the given generic function.  If \code{gf} is not supplied, all
  cached effective methods are flushed.
\end{defun}

\begin{defvar}{pcl::}{inline-methods-in-emfs}
  If true, the default, perform method inlining as described above.
  If false, don't.
\end{defvar}



\subsection{Effective Method Precomputation}
\cpsubindex{effective method}{precomputation}
\cpsubindex{methods}{load time}
\cpsubindex{methods}{emf precomputation}

When a generic function is called, the generic function's
discriminating function computes the set of methods applicable to
actual arguments and constructs an effective method function from
applicable methods, using the generic function's method combination.

Effective methods can be precomputed at method load time instead of
when the generic function is called depending on the value of
\code{pcl:*max-emf-precomputation-methods*}.

\begin{defvar}{pcl:}{*max-emf-precomputation-methods*}
  If nonzero, the default value is 100, precompute effective methods
  when methods are loaded, and the method's generic function has less
  than the specified number of methods.
  
  If zero, compute effective methods only when the generic function is
  called.
\end{defvar}



\subsection{Sealing}
\cindex{sealing}
\cpsubindex{sealing}{subclasses}
\cpsubindex{sealing}{methods}
\cpsubindex{methods}{sealing}

Support for sealing classes and generic functions have been
implemented.  Please note that this interface is subject to change.

\begin{defmac}{pcl:}{seal}{\var{name} (\var{var}) \amprest{} \var{specifiers}}
  Seal \code{name} with respect to the given specifiers; \code{name}
  can be the name of a class or generic-function.

  Supported specifiers are \kwd{subclasses} for classes,
  which prevents changing subclasses of a class, and \kwd{methods}
  which prevents changing the methods of a generic function.
  
  Sealing violations signal an error of type \code{pcl:sealed-error}.
\end{defmac}

\begin{defun}{pcl:}{unseal}{\var{name-or-object}}
  Remove seals from \code{name-or-object}.
\end{defun}



\subsection{Method Tracing and Profiling}
\label{sec:method-tracing}
\cindex{tracing}
\cpsubindex{tracing}{methods}
\cindex{profiling}
\cpsubindex{profiling}{methods}
\cpsubindex{methods}{tracing}
\cpsubindex{methods}{profiling}

Methods can be traced with \code{trace}, using function names of the
form \code{(method <name> <qualifiers> <specializers>)}.  Example:

\begin{example}
(defmethod foo ((x integer)) x)
(defmethod foo :before ((x integer)) x)

(trace (method foo (integer)))
(trace (method foo :before (integer)))
(untrace (method foo :before (integer)))
\end{example}
  
\code{trace} and \code{untrace} also allow a name specifier
\code{:methods gf-form} for tracing all methods of a generic function:

\begin{example}
(trace :methods 'foo)
(untrace :methods 'foo)
\end{example}

Methods can also be specified for the \kwd{wherein} option to
\code{trace}.  Because this option is a name or a list of names,
methods must be specified as a list.  Thus, to trace all calls of
\code{foo} from the method \code{bar} specialized on integer argument,
use
\begin{example}
  (trace foo :wherein ((method bar (integer))))
\end{example}
Before and after methods are supported as well:
\begin{example}
  (trace foo :wherein ((method bar :before (integer))))
\end{example}

Method profiling is done analogously to \code{trace}:

\begin{example}
(defmethod foo ((x integer)) x)
(defmethod foo :before ((x integer)) x)

(profile:profile (method foo (integer)))
(profile:profile (method foo :before (integer)))
(profile:unprofile (method foo :before (integer)))

(profile:profile :methods 'foo)
(profile:unprofile :methods 'foo)

(profile:profile-all :methods t)
\end{example}



\subsection{Misc}
\cpsubindex{methods}{interpreted}

\begin{defvar}{pcl::}{compile-interpreted-methods-p}
  This variable controls compilation of interpreted method functions,
  e.g. for methods defined interactively at the REPL.  Default is
  true, that is, method functions are compiled.
\end{defvar}





\section{Differences from ANSI Common Lisp}
This section describes some of the known differences between \cmucl{}
and ANSI \clisp{}.  Some may be non-compliance issues; same may be
extensions.

\subsection{Extensions}

\begin{defun}{}{constantly}{%
    \var{value} \ampoptional{} \var{val1} \var{val2} \amprest{} \var{more-values}}
  As an extension, \cmucl{} allows \code{constantly} to accept more
  than one value which are returned as multiple values.
\end{defun}




\section{Function Wrappers}
\cindex{function wrappers}
\cindex{fwrappers}

Function wrappers, fwrappers for short, are a facility for efficiently
encapsulating functions\footnote{This feature was independently
developed, but the interface is modelled after a similar feature in
Allegro.  Some names, however, have been changed.}.

Functions in \cmucl{} are represented by \code{kernel:fdefn}
objects.  Each \code{fdefn} object contains a reference to its
function's actual code, which we call the function's primary function.

A function wrapper replaces the primary function in the \code{fdefn}
object with a function of its own, and records the original function
in an fwrapper object, a funcallable instance.  Thus, when the
function is called, the fwrapper gets called, which in turn might call
the primary function, or a previously installed fwrapper that was
found in the \code{fdefn} object when the second fwrapper was
installed.

Example:

\begin{lisp}
(use-package :fwrappers)

(define-fwrapper foo (x y)
  (format t "x = ~s, y = ~s, user-data = ~s~%"
          x y (fwrapper-user-data fwrapper))
  (let ((value (call-next-function)))
    (format t "value = ~s~%" value)
    value))

(defun bar (x y)
  (+ x y))

(fwrap 'bar #'foo :type 'foo :user-data 42)

(bar 1 2)
 =>
 x = 1, y = 2, user-data = 42
 value = 3
 3   
\end{lisp}

Fwrappers are used in the implementation of \code{trace} and
\code{profile}.

Please note that \code{fdefinition} always returns the primary
definition of a function; if a function is fwrapped,
\code{fdefinition} returns the primary function stored in the
innermost fwrapper object.  Likewise, if a function is fwrapped,
\code{(setf fdefinition)} will set the primary function in the
innermost fwrapper.

\begin{defmac}{fwrappers:}{define-fwrapper}{\var{name} \var{lambda-list} \ampbody{} \var{body}}
  This macro is like \code{defun}, but defines a function named
  \var{name} that can be used as an fwrapper definition.
  
  In \var{body}, the symbol \code{fwrapper} is bound to the current
  fwrapper object.
  
  The macro \code{call-next-function} can be used to invoke the next
  fwrapper, or the primary function that is being fwrapped.  When
  called with no arguments, \code{call-next-function} invokes the next
  function with the original arguments passed to the fwrapper, unless
  you modify one of the parameters.  When called with arguments,
  \code{call-next-function} invokes the next function with the given
  arguments.
\end{defmac}

\begin{defun}{fwrappers:}{fwrap}{\var{function-name} \var{fwrapper} %
    \keys{\kwd{type} \kwd{user-data}}}
  This function wraps function \code{function-name} in an fwrapper
  \var{fwrapper} which was defined with \code{define-fwrapper}.

  The value of \var{type}, if supplied, is used as an identifying
  tag that can be used in various other operations.
  
  The value of \var{user-data} is stored as user-supplied data in the
  fwrapper object that is created for the function encapsulation.
  User-data is accessible in the body of fwrappers defined with
  \code{define-fwrapper} as \code{(fwrapper-user-data fwrapper)}.

  Value is the fwrapper object created.
\end{defun}

\begin{defun}{fwrappers:}{funwrap}{\var{function-name} \keys{\kwd{type} \kwd{test}}}
  Remove fwrappers from the function named \var{function-name}.  If
  \var{type} is supplied, remove fwrappers whose type is \code{equal}
  to \var{type}.  If \var{test} is supplied, remove fwrappers
  satisfying \var{test}.
\end{defun}

\begin{defun}{fwrappers:}{find-fwrapper}{\var{function-name} \keys{\kwd{type} \kwd{test}}}
  Find an fwrapper of \var{function-name}.  If \var{type} is supplied,
  find an fwrapper whose type is \code{equal} to \var{type}.  If
  \var{test} is supplied, find an fwrapper satisfying \var{test}.
\end{defun}

\begin{defun}{fwrappers:}{update-fwrapper}{\var{fwrapper}}
  Update the funcallable instance function of the fwrapper object
  \var{fwrapper} from the definition of its function that was 
  defined with \code{define-fwrapper}.  This can be used to update
  fwrappers after changing a \code{define-fwrapper}.
\end{defun}

\begin{defun}{fwrappers:}{update-fwrappers}{\var{function-name} \keys{\kwd{type} \kwd{test}}}
  Update fwrappers of \var{function-name}; see \code{update-fwrapper}.
  If \var{type} is supplied, update fwrappers whose type is
  \code{equal} to \var{type}.  If \var{test} is supplied, update fwrappers
  satisfying \var{test}.
\end{defun}

\begin{defun}{fwrappers:}{set-fwrappers}{\var{function-name} \var{fwrappers}}
  Set \var{function-names}'s fwrappers to elements of the list
  \var{fwrappers}, which is assumed to be ordered from outermost to
  innermost.  \var{fwrappers} null means remove all fwrappers.
\end{defun}

\begin{defun}{fwrappers:}{list-fwrappers}{\var{function-name}}
  Return a list of all fwrappers of \var{function-name}, ordered
  from outermost to innermost.
\end{defun}

\begin{defun}{fwrappers:}{push-fwrapper}{\var{fwrapper} \var{function-name}}
  Prepend fwrapper \var{fwrapper} to the definition of
  \var{function-name}.  Signal an error if \var{function-name} is an
  undefined function.
\end{defun}

\begin{defun}{fwrappers:}{delete-fwrapper}{\var{fwrapper} \var{function-name}}
  Remove fwrapper \var{fwrapper} from the definition of
  \var{function-name}.  Signal an error if \var{function-name} is an
  undefined function.
\end{defun}

\begin{defmac}{fwrappers:}{do-fwrappers}{(\var{var} \var{fdefn} \ampoptional{}
  \var{result}) \ampbody{} \var{body}}
  Evaluate \var{body} with \var{var} bound to consecutive fwrappers of
  \var{fdefn}.  Return \var{result} at the end.  Note that \var{fdefn}
  must be an \code{fdefn} object.  You can use
  \code{kernel:fdefn-or-lose}, for instance, to get the \code{fdefn}
  object from a function name.
\end{defmac}

\section{Dynamic-Extent Declarations}
\cindex{dynamic-extent}

\emph{Note:  As of the 19a release, \code{dynamic-extent} is
  unfortunately disabled by default.  It is known to cause some issues
  with CLX and Hemlock.  The cause is not known, but causes random
  errors and brokeness.  Enable at your own risk.  However, it is safe
  enough to build all of CMUCL without problems.}

On x86 and sparc, \cmucl{} can exploit \code{dynamic-extent}
declarations by allocating objects on the stack instead of the heap.

You can tell \cmucl{} to trust or not trust \code{dynamic-extent}
declarations by setting the variable
\var{*trust-dynamic-extent-declarations*}.

\begin{defvar}{ext:}{trust-dynamic-extent-declarations}
  If the value of \var{*trust-dynamic-extent-declarations*} is 
  \code{NIL}, \code{dynamic-extent} declarations are effectively
  ignored.

  If the value of this variable is a function, the function is called
  with four arguments to determine if a \code{dynamic-extent} 
  declaration should be trusted.  The arguments are the safety,
  space, speed, and debug settings at the point where the 
  \code{dynamic-extent} declaration is used.  If the function
  returns true, the declaration is trusted, otherwise it is not
  trusted.

  In all other cases, \code{dynamic-extent} declarations are
  trusted.
\end{defvar}

Please note that stack-allocation is inherently unsafe.  If you make a
mistake, and a stack-allocated object or part of it escapes, \cmucl{}
is likely to crash, or format your hard disk.

\subsection{\code{\&rest} argument lists}
\cpsubindex{dynamic-extent}{rest lists}

Rest argument lists can be allocated on the stack by declaring the
rest argument variable \code{dynamic-extent}.  Examples:

\begin{lisp}
(defun foo (x &rest rest)
  (declare (dynamic-extent rest))
  ...)

(defun bar ()
  (lambda (&rest rest)
    (declare (dynamic-extent rest))
    ...))
\end{lisp}

\subsection{Closures}
\cpsubindex{dynamic-extent}{closures}

Closures for local functions can be allocated on the stack if the
local function is declared \code{dynamic-extent}, and the closure
appears as an argument in the call of a named function.  In the
example:

\begin{lisp}
(defun foo (x)
  (flet ((bar () x))
    (declare (dynamic-extent #'bar))
    (baz #'bar)))
\end{lisp}

the closure passed to function \code{baz} is allocated on the stack.
Likewise in the example:

\begin{lisp}
(defun foo (x)
  (flet ((bar () x))
    (baz #'bar)
    (locally (declare (dynamic-extent #'bar))
      (baz #'bar))))
\end{lisp}

\cpsubindex{dynamic-extent}{known CL functions}

Stack-allocation of closures can also automatically take place when
calling certain known CL functions taking function arguments, for
example \code{some} or \code{find-if}.

\subsection{\code{list}, \code{list*}, and \code{cons}}
\cpsubindex{dynamic-extent}{list, list*, cons}

New conses allocated by \code{list}, \code{list*}, or \code{cons}
which are used to initialize variables can be allocated from the stack
if the variables are declared \code{dynamic-extent}.  In the case of
\code{cons}, only the outermost cons cell is allocated from the stack;
this is an arbitrary restriction.

\begin{lisp}
(let ((x (list 1 2))
      (y (list* 1 2 x))
      (z (cons 1 (cons 2 nil))))
  (declare (dynamic-extent x y z))
  ...
  (setq x (list 2 3))
  ...)
\end{lisp}

Please note that the \code{setq} of \code{x} in the example program
assigns to \code{x} a list that is allocated from the heap.  This is
another arbitrary restriction that exists because other Lisps behave
that way.

\section{Modular Arithmetic}
\cindex{modular-arith}

This section is mostly taken, with permission,  from the documentation
for SBCL.

Some numeric functions have a property: \code{N} lower bits of
the result depend only on \code{N} lower bits of (all or some)
arguments. If the compiler sees an expression of form \code{(logand
exp mask)}, where \code{exp} is a tree of such ``good'' functions
and \code{mask} is known to be of type \code{(unsigned-byte
w)}, where \code{w} is a "good" width, all intermediate results
will be cut to \code{w} bits (but it is not done for variables
and constants!). This often results in an ability to use simple
machine instructions for the functions.

Consider an example.
\begin{lisp}
(defun i (x y)
  (declare (type (unsigned-byte 32) x y))
  (ldb (byte 32 0) (logxor x (lognot y))))
\end{lisp}
The result of \code{(lognot y)} will be negative and of
type \code{(signed-byte 33)}, so a naive implementation on a 32-bit
platform is unable to use 32-bit arithmetic here. But modular
arithmetic optimizer is able to do it: because the result is cut down
to 32 bits, the compiler will replace \code{logxor}
and \code{lognot} with versions cutting results to 32 bits, and
because terminals (here---expressions \code{x} and \code{y})
are also of type \code{(unsigned-byte 32)}, 32-bit machine
arithmetic can be used.


Currently ``good'' functions
are \code{+}, \code{-}, \code{*}; \code{logand}, \code{logior},
\code{logxor}, \code{lognot} and their combinations;
and \code{ash} with the positive second argument. ``Good'' widths
are 32 on HPPA, MIPS, PPC, Sparc and X86 and 64 on Alpha. While it is
possible to support smaller widths as well, currently it is not
implemented.

A more extensive description of modular arithmetic can be found in the
paper ``Efficient Hardware Arithmetic in Common Lisp'' by Alexey
Dejneka, and Christophe Rhodes, to be published.

\section{Extension to REQUIRE}
\cindex{require}

The behavior of \code{require} when called with only one argument is
implementation-defined.  In \cmucl, functions from the list
\var{*module-provider-functions*} are called in order with the
stringified module name as the argument.  The first function to return
non-\var{NIL} is assumed to have loaded the module.

By default the functions \code{module-provide-cmucl-defmodule} and
\code{module-provide- cmucl-library} are on this list of functions, in
that order.

\begin{defvar}{ext:}{module-provider-functions}
  This is a list of functions taking a single argument.
  \code{require} calls each function in turn with the stringified
  module name.  The first function to return non-\var{NIL} indicates
  that the module has been loaded.  The remaining functions, if any,
  are not called.

  To add new providers, push the new provider function onto the
  beginning of this list.
\end{defvar}

\begin{defmac}{ext:}{defmodule}{\var{name} \amprest{} \var{files}}
  Defines a module by registering the files that need to be loaded
  when the module is required.  If \var{name} is a symbol, its print
  name is used after downcasing it.
\end{defmac}

\begin{defun}{ext:}{module-provide-cmucl-defmodule}{\var{module-name}}
  This function is the module-provider for modules registered by a
  \code{ext:defmodule} form.  
\end{defun}

\begin{defun}{ext:}{module-provide-cmucl-library}{\var{module-name}}
  This function is the module-provider for \cmucl's libraries,
  including Gray streams, simple streams, CLX, CLM, Hemlock,
  \emph{etc}.
  
  This function causes a file to be loaded whose name is formed by
  merging the search-list ``modules:'' and the concatenation of
  module-name with the suffix ``-LIBRARY''.  Note that both the
  module-name and the suffix are each, separately, converted from
  :case :common to :case :local.  This merged name will be probed with
  both a .lisp and .fasl extensions, calling \code{LOAD} if it exists.
\end{defun}


\section{Localization}
\label{sec:localization}

\cmucl{} support localization where messages can be presented in the
native language.  This is done in the style of \code{gettext} which
marks strings that are to be translated and provides the lookup to
convert the string to the specified language.

All messages from \cmucl{} can be translated but as of this writing,
the only complete translation is a Pig Latin translation done by
machine.  There are a few messages translated to Korean.

In general, translatable strings are marked as such by using the
functions \code{intl:gettext} and \code{intl:ngettext} or by using the
reader macros \verb+_+ or \verb+_N+.  When loading or compiling, such
strings are recorded for translation.  At runtime, such strings are
looked in and the translation is returned.  Doc strings do not need to
be noted in any way; the are automatically noted for translation.

By default, recording of translatable strings is disabled.  To enable
recording of strings, call \code{intl:translation-enable}.

\subsection{Dictionary}
\label{sec:localization-dictionary}

\begin{defun}{intl:}{translation-enable}{}
  Enable recording of translatable strings.
\end{defun}

\begin{defun}{intl:}{translation-disable}{}
  Disablle recording of translatable strings.
\end{defun}

\begin{defun}{intl:}{setlocale}{\ampoptional{} \var{locale}}
  Sets the locale to the locale specified by \var{locale}.  If
  \var{locale} is not give or is \nil, the locale is determined by
  look at the environment variables \code{LANGUAGE}, \code{LC\_ALL},
  \code{LC\_MESSAGES}, or \code{LANG}.  If none of these are set, the
  locale is unchanged.

  The default locale is ``C''.
\end{defun}

\begin{defun}{intl:}{textdomain}{\var{domain}}
  Set the default domain to the domain specified by \var{domain}.
  Typically,  this only needs to be done at the top of each source
  file.  This is used to \code{gettext} and \code{ngettext} to set the
  domain for the message string.
\end{defun}

\begin{defmac}{intl:}{gettext}{\var{string}}
  Look up the specified string, \var{string}, in the current message
  domain and return its translation.
\end{defmac}

\begin{defun}{intl:}{dgettext}{\var{domain} \var{string}}
  Look up the specified string, \var{string}, in the message domain,
  \var{domain}.  The translation is returned.

  When compiled, this also function also records the string so that an
  appropriate message template file can be created.  (See
  \code{intl::dump-pot-files}.) 
\end{defun}

\begin{defmac}{intl:}{ngettext}{\var{singular} \var{plural} \var{n}}
  Look up the singular or plural form of a message in the default
  domain.  The singular form is \var{singular}; the plural is
  \var{plural}.  The number of items is specified by \var{n} in case
  the correct translation depends on the actual number of items.
\end{defmac}

\begin{defun}{intl:}{dngettext}{\var{domain} \var{singular} \var{plural} \var{n}}
  Look up the singular or plural form of a message in the specified
  domain, \var{domain}.  The singular form is \var{singular}; the
  plural is \var{plural}.  The number of items is specified by \var{n}
  in case the correct translation depends on the actual number of
  items.

  When compiled, this also function also records the singular and
  plural forms so that an appropriate message template file can be
  created.  (See \code{intl::dump-pot-files}.)
\end{defun}

\begin{defun}{intl::}{dump-pot-files}{\keys{\kwd{copyright} \kwd{output-directory}}}
  Dumps the translatable strings recorded by \code{dgettext} and
  \code{dngettext}.  The message template file (pot file) is written
  to a file in the directory specified by \var{output-directory}, and
  the name of the file is the domain of the string.

  If \var{copyright} is specified, this is placed in the output file
  as the copyright message.
\end{defun}

\begin{defvar}{intl:}{locale-directories}
  This is a list of directory pathnames where the translations can be found.
\end{defvar}  

\begin{defun}{intl:}{install}{\ampoptional{} (\var{rt} \var{*readtable*})}
  Installs reader macros and comment reader into the specified
  readtable as explained below.  The readtable defaults to
  \var{*readtable*}.
\end{defun}

Two reader macros are also provided: \code{\_''} and \code{\_N''}.  The
first is equivalent to wrapping \code{dgettext} around the string.
The second returns the string, but also records the string.  This is
needed when we want to record a docstring for translation or any other
string in a place where a macro or function call would be incorrect.

Also, the standard comment reader is extended to allow translator
comments to be saved and written to the messages template file so that
the translator may not need to look at the original source to
understand the string.  Any comment line that begins with exactly
\verb|"TRANSLATORS: "| is saved.  This means each translator comment
must be preceded by this string to be saved; the translator comment
ends at the end of each line.


\subsection{Example Usage}
\label{sec:localization-usage}

Here is a simple example of how to localize your code.  Let the file
\code{intl-ex.lisp} contain:

\begin{example}

(intl:textdomain "example")  

(defun foo (x y)
  "Cool function foo of x and y"
  (let ((result (bar x y)))
    ;; TRANSLATORS:  One line comment about bar.
    (format t _"bar of ~A and ~A = ~A~%" x y result)
    #| TRANSLATORS:  Multiline comment about
    how many Xs there are
    |#
    (format t (intl:ngettext "There is one X"
                             "There are many Xs"
                             x))
    result))
\end{example}

The call to \code{textdomain} sets the default domain for all
translatable strings following the call.

Here is a sample session for creating a template file:

\begin{example}
* (intl:install)

T
* (intl:translation-enable)

T
* (compile-file "intl-ex")

#P"/Volumes/share/cmucl/cvs/intl-ex.sse2f"
NIL
NIL
* (intl::dump-pot-files :output-directory "./")

Dumping 3 messages for domain "example"
NIL
*
\end{example}

When this file is compiled, all of the translatable strings are
recorded.  This includes the docstring for \code{foo}, the string for
the first \code{format}, and the string marked by the call to
\code{intl:ngettext}.

A file named ``example.pot'' in the directory ``./'' is created.
The contents of this file are:
\begin{example}
#@ example

# SOME DESCRIPTIVE TITLE
# FIRST AUTHOR <EMAIL@ADDRESS>, YEAR
#
#, fuzzy
msgid ""
msgstr ""
"Project-Id-Version: PACKAGE VERSION"
"Report-Msgid-Bugs-To: "
"PO-Revision-Date: YEAR-MO-DA HO:MI +ZONE"
"Last-Translator: FULL NAME <EMAIL@ADDRESS>"
"Language-Team: LANGUAGE <LL@li.org>"
"MIME-Version: 1.0"
"Content-Type: text/plain; charset=UTF-8"
"Content-Transfer-Encoding: 8bit"

#.  One line comment about bar.
#: intl-ex.lisp
msgid "bar of ~A and ~A = ~A~%"
msgstr ""

#.  Multiline comment about
    how many Xs there are
#: intl-ex.lisp
msgid "Cool function foo of x and y"
msgstr ""

#: intl-ex.lisp
msgid "There is one X"
msgid_plural "There are many Xs"
msgstr[0] ""

\end{example}

To finish the translation, a corresponding ``example.po'' file needs
to be created with the appropriate translations for the given
strings.  This file must be placed in some directory that is included
in \code{intl:*locale-directories*}.

Suppose the translation is done for Korean.  Then the user can set the
environment variables appropriately or call \code{(intl:setlocale
  "ko")}.  Note that the external format for the standard streams
needs to be set up appropriately too.  It is up to the user to set
this correctly.  Once this is all done, the output from the function
\code{foo} will now be in Korean instead of English as in the original
source file.

For further information, we refer the reader to documentation on
\ifpdf
\href{http://www.gnu.org/software/gettext/manual/gettext.html}{gettext}.
\else
gettext at
\href{http://www.gnu.org/software/gettext/manual/gettext.html}{\texttt{http://www.gnu.org/software/gettext/manual/gettext.html}}.
\fi

\section{Static Arrays}
\label{sec:static-arrays}

\cmucl{} supports static arrays which are arrays that are not moved by
the garbage collector.  To create such an array, use the
\kwd{allocation} option to \code{make-array} with a value of
\kwd{malloc}.  These arrays appear as normal Lisp arrays, but are
actually allocated from the \code{C} heap (hence the \kwd{malloc}).
Thus, the number and size of such arrays are limited by the available
\code{C} heap.

Also, only certain types of arrays can be allocated.  The static array
cannot be adjustable and cannot be displaced to.  The array must also
be a \code{simple-array} of one dimension.  The element type is also
constrained to be one of the types in
Table~\ref{tbl:static-array-types}.

\begin{table}[tbhp]
  \begin{center}
    \begin{tabular}{|c|}
      \hline
      \code{(unsigned-byte 8)} \\
      \hline
      \code{(unsigned-byte 16)} \\
      \hline
      \code{(unsigned-byte 32)} \\
      \hline
      \code{(signed-byte 8)} \\
      \hline
      \code{(signed-byte 16)} \\
      \hline
      \code{(signed-byte 32)} \\
      \hline
      \code{single-float} \\
      \hline
      \code{double-float} \\
      \hline
      \code{(complex single-float)} \\
      \hline
      \code{(complex double-float)} \\
      \hline
    \end{tabular}
    \caption{Allowed element types for static arrays}
    \label{tbl:static-array-types}
  \end{center}
\end{table}

The arrays are properly handled by GC.  GC will not move the arrays,
but they will be properly removed up if they become garbage.
